---
title: "Data Codebook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Codebook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r general_setup, include=FALSE, message=FALSE}
library(parkinsonsMetagenomicData)
library(dplyr)
library(knitr)
```

# Data Codebook

This vignette documents all variables and data structures in the `parkinsonsMetagenomicData` package, including sample metadata fields, microbiome data types, and column definitions for parquet files.

## Overview

The package provides three main data components:

1. **Sample Metadata** (`sampleMetadata`) - Clinical and demographic information for each sample
2. **Microbiome Data** - Multiple data types from MetaPhlAn, HUMAnN, and QC tools
3. **Reference Files** - Lookup tables for taxonomic names, gene families, and pathways

## Sample Metadata

The `sampleMetadata` data frame contains curated metadata for all samples. Both curated and uncurated features are included, with uncurated features prefixed by "uncurated_".

### Loading Sample Metadata

```{r load_metadata}
data("sampleMetadata", package = "parkinsonsMetagenomicData")
dim(sampleMetadata)
```

### Core Metadata Fields

#### Identifiers

| Field | Type | Description |
|-------|------|-------------|
| `curation_id` | character | Dataset x subject identifier (format: `study_name:subject_id`) |
| `study_name` | character | Dataset name |
| `subject_id` | character | Subject identifier within study |
| `sample_id` | character | Unique sample identifier |
| `uuid` | character | Universal unique identifier for the sample |
| `BioProject` | character | SRA BioProject accession (format: PRJ[DEN][BA][0-9]+) |
| `BioSample` | character | SRA BioSample accession (format: SAM[DNEA]+?[0-9]+) |
| `NCBI_accession` | character | Semicolon-separated vector of NCBI accessions |

#### Study Design

| Field | Type | Description | Allowed Values |
|-------|------|-------------|----------------|
| `target_condition` | character | Primary phenotype/condition of interest (multiple values separated by `;`) | Ontology terms (NCIT:C7057, EFO:0000408 descendants) |
| `control` | character | Sample classification in study | "Study Control", "Case", "Not Used" |

#### Demographics

| Field | Type | Description | Allowed Values |
|-------|------|-------------|----------------|
| `age` | integer | Age of subject | Numeric value |
| `age_unit` | character | Unit for age | "Day", "Week", "Month", "Year" |
| `age_group` | character | Age category | "Infant" (0-2), "Children 2-11 Years Old" (2-11), "Adolescent" (11-18), "Adult" (18-65), "Elderly" (≥65) |
| `sex` | character | Biological sex | "Female", "Male" |
| `host_species` | character | Species of subject | "Homo sapiens", "Mus musculus" |

#### Clinical Information

| Field | Type | Description | Allowed Values |
|-------|------|-------------|----------------|
| `disease` | character | Reported disease/condition(s) (multiple values separated by `;`; "Healthy" if none) | Ontology terms (NCIT:C7057, EFO:0000408 descendants) |
| `body_site` | character | Anatomical location | "feces", "milk", "nasal cavity", "oral cavity", "skin epidermis", "vagina" |

#### Curation

| Field | Type | Description |
|-------|------|-------------|
| `curator` | character | Curator name(s) (multiple values separated by `;`) |

### Accessing Metadata Information Programmatically

```{r metadata_info}
# View the data dictionary
metadata_info <- data_dict()
kable(head(metadata_info, 10))
```

### Example: Exploring Sample Characteristics

```{r explore_metadata}
# Summary of control status
table(sampleMetadata$control, useNA = "ifany")

# Age distribution
summary(sampleMetadata$age)

# Studies included
head(unique(sampleMetadata$study_name))

# Body sites sampled
table(sampleMetadata$body_site, useNA = "ifany")
```

## Microbiome Data Types

The package provides multiple types of microbiome profiling data, organized by biological content and normalization method.

### Available Data Types

```{r data_types}
# Get information about all data types
data_types <- biobakery_files()
kable(data_types)
```

### Data Type Categories

#### Taxonomic Composition (MetaPhlAn)

**`relative_abundance`** - Primary taxonomic profiling output

- **Description**: Relative abundance of bacterial, archaeal, viral, and eukaryotic taxa
- **Units**: Relative Abundance (%)
- **Levels**: Kingdom through species and strain
- **Tool**: MetaPhlAn

**`viral_clusters`** - Viral community profiling

- **Description**: Clusters of viral sequences for strain/species-level analysis
- **Units**: Varies (often Relative Abundance or CPM)
- **Tool**: MetaPhlAn/Custom

**`marker_abundance`** - Marker gene quantification

- **Description**: Abundance of clade-specific marker genes
- **Units**: Mean coverage of marker genes
- **Tool**: MetaPhlAn

**`marker_presence`** - Marker gene detection

- **Description**: Binary presence/absence of marker genes
- **Units**: Binary (0 or 1)
- **Tool**: MetaPhlAn

#### Strain-Level Profiling (StrainPhlAn)

**`strainphlan_markers`** - Strain-specific markers

- **Description**: Consensus sequences for strain-level reconstruction
- **Units**: Fraction of marker covered / Mean coverage
- **Tool**: StrainPhlAn

#### Functional Profiling (HUMAnN)

HUMAnN data types come in multiple variants based on:
- **Content**: Gene families (`genefamilies`), pathway abundance (`pathabundance`), or pathway coverage (`pathcoverage`)
- **Stratification**: Taxonomically stratified (by species) or unstratified (community total)
- **Normalization**: Raw (RPK), relative abundance (`relab`), or copies per million (`cpm`)

**Gene Families**

| Data Type | Stratification | Normalization | Description |
|-----------|----------------|---------------|-------------|
| `genefamilies` | Mixed | RPK | Raw output (unnormalized) |
| `genefamilies_unstratified` | Community total | RPK | Total gene family abundance |
| `genefamilies_stratified` | By species | RPK | Species-specific contributions |
| `genefamilies_relab_unstratified` | Community total | Relative % | Normalized community totals |
| `genefamilies_relab_stratified` | By species | Relative % | Normalized species contributions |
| `genefamilies_cpm_unstratified` | Community total | CPM | Depth-corrected totals |
| `genefamilies_cpm_stratified` | By species | CPM | Depth-corrected species contributions |

**Metabolic Pathways - Abundance**

| Data Type | Stratification | Normalization | Description |
|-----------|----------------|---------------|-------------|
| `pathabundance` | Mixed | RPK | Raw pathway abundance |
| `pathabundance_unstratified` | Community total | RPK | Total pathway abundance |
| `pathabundance_stratified` | By species | RPK | Species-specific pathway contributions |
| `pathabundance_relab_unstratified` | Community total | Relative % | Normalized community pathways |
| `pathabundance_relab_stratified` | By species | Relative % | Normalized species pathways |
| `pathabundance_cpm_unstratified` | Community total | CPM | Depth-corrected pathway totals |
| `pathabundance_cpm_stratified` | By species | CPM | Depth-corrected species pathways |

**Metabolic Pathways - Coverage**

| Data Type | Stratification | Units | Description |
|-----------|----------------|-------|-------------|
| `pathcoverage` | Mixed | Proportion (0-1) | Pathway completeness |
| `pathcoverage_unstratified` | Community total | Proportion (0-1) | Total pathway coverage |
| `pathcoverage_stratified` | By species | Proportion (0-1) | Species-specific pathway coverage |

#### Quality Control

**`fastqc`** - Sequencing quality metrics

- **Description**: Quality scores, adapter contamination, sequence content
- **Units**: Varies by metric
- **Tool**: FastQC

**`kneaddata_log`** - Preprocessing statistics

- **Description**: Quality trimming, host read removal, read counts
- **Units**: None (log file)
- **Tool**: KneadData

### Choosing Data Types

**For taxonomic analysis:**
- Use `relative_abundance` for bacteria/archaea
- Add `viral_clusters` for viruses

**For functional analysis:**
- Use `genefamilies_relab` or `pathabundance_relab` for relative comparisons
- Use `genefamilies_cpm` or `pathabundance_cpm` when comparing across sequencing depths
- Use `pathcoverage` to assess pathway completeness

**Stratified vs. Unstratified:**
- Unstratified: Total community-level measurements (smaller files)
- Stratified: See which species contribute to each function (larger files, richer information)

## Parquet File Column Definitions

Each data type has specific columns with defined roles in the TreeSummarizedExperiment structure.

### Column Roles

Columns are assigned to specific components:

- **`cname`**: Column names (sample identifiers, typically `uuid`)
- **`cdata`**: Column data (sample metadata: processing parameters, versions)
- **`rname`**: Row names (feature identifiers)
- **`rdata`**: Row data (feature metadata)
- **`assay`**: Assay data (measurement values)

### Accessing Column Information

```{r column_info}
# Get column information for a specific data type
rel_abund_cols <- parquet_colinfo("relative_abundance")
kable(rel_abund_cols)
```

### Relative Abundance Columns

```{r relab_cols}
# Show relative abundance columns grouped by role
rel_abund_cols %>%
    select(col_name, col_class, se_role, description) %>%
    kable()
```

#### Sample Identifiers (cname)
- **`uuid`**: Sample UUID - links to `sampleMetadata`

#### Sample Metadata (cdata)
- **`db_version`**: MetaPhlAn database version
- **`command`**: MetaPhlAn command executed
- **`reads_processed`**: Number of reads processed
- **`metaphlan_header`**: Original MetaPhlAn header
- **`original_columns`**: Original column names

#### Feature Identifiers (rname)
- **`clade_name`**: Full taxonomic lineage (e.g., "k__Bacteria|p__Firmicutes|c__Bacilli|o__Lactobacillales|f__Streptococcaceae|g__Streptococcus|s__Streptococcus_mutans")

#### Feature Metadata (rdata)
- **`clade_name_kingdom`** through **`clade_name_terminal`**: Taxonomic ranks parsed from `clade_name`
- **`NCBI_tax_id`** through **`NCBI_tax_id_terminal`**: NCBI Taxonomy IDs for each rank
- **`additional_species`**: Other species with the same marker set

#### Assay Data
- **`relative_abundance`**: Proportion of community (0-100%)

### Gene Families Columns

```{r genefamilies_cols}
# Gene families column structure
gf_cols <- parquet_colinfo("genefamilies")
gf_cols %>%
    select(col_name, col_class, se_role, description) %>%
    kable()
```

Key columns:
- **`gene_family`** (rname): Full gene family identifier
- **`gene_family_uniref`** (rdata): UniRef90 identifier
- **`gene_family_genus`** (rdata): Taxonomic genus (if stratified)
- **`gene_family_species`** (rdata): Taxonomic species (if stratified)
- **`rpk_abundance`** (assay): Reads per kilobase

### Pathway Columns

```{r pathway_cols}
# Pathway abundance/coverage column structure
pa_cols <- parquet_colinfo("pathabundance")
pa_cols %>%
    select(col_name, col_class, se_role, description) %>%
    kable()
```

Key columns:
- **`pathway`** (rname): MetaCyc pathway identifier
- **`pathway_uniref`** (rdata): UniRef identifier (if applicable)
- **`pathway_genus`** (rdata): Taxonomic genus (if stratified)
- **`pathway_species`** (rdata): Taxonomic species (if stratified)
- **`abundance`** or **`coverage`** (assay): Measurement value

## Reference Files

Reference files provide lookup tables for non-UUID identifiers found in the parquet files.

### Available References

```{r ref_info}
# Get information about reference files
ref_info <- get_ref_info()
kable(ref_info)
```

### Loading Reference Files

Reference files can be loaded from remote repositories or local files.

```{r load_ref_example, eval=FALSE}
# From remote repository (requires network)
clade_ref <- load_ref("clade_name_ref")

# From local file
refpath <- file.path(system.file("extdata",
                                 package = "parkinsonsMetagenomicData"),
                     "pathway_ref.parquet")
pathway_ref <- load_ref(ref_file = refpath)
head(pathway_ref)
```

### Reference File Contents

**`clade_name_ref`** - Taxonomic lineages

- All unique `clade_name` values from `relative_abundance` files
- Includes NCBI Taxonomy IDs at each rank
- Use for: Taxonomic lookups, lineage parsing

**`gene_family_ref`** - Gene family identifiers

- All unique `gene_family` values from `genefamilies` files
- Use for: Gene family annotation lookups

**`pathway_ref`** - Pathway identifiers

- All unique `pathway` values from `pathabundance` and `pathcoverage` files
- Use for: Pathway annotation, MetaCyc lookups

**`genome_name_ref`** - Viral genome identifiers

- All unique `genome_name` values from `viral_clusters` files
- Use for: Viral genome annotation

**`uniref_marker_ref`** - Marker gene identifiers

- All unique `uniref` marker values from `marker_abundance` and `marker_presence` files
- Use for: Marker gene lookups

## Data Repositories

Data are hosted on Hugging Face in parquet format for efficient access.

### Repository Information

```{r repo_info}
# Get repository information
repos <- get_repo_info()
kable(repos)
```

### Default Repository

The default repository (`waldronlab/metagenomics_mac`) contains the full dataset with all available samples.

### Examples Repository

The examples repository (`waldronlab/metagenomics_mac_examples`) contains small example files with 10 samples each, useful for testing and learning.

## Data Access Patterns

### Quick Access with returnSamples()

For most users, `returnSamples()` provides the simplest interface:

```{r return_samples_example, eval=FALSE}
# Load sample metadata
data("sampleMetadata", package = "parkinsonsMetagenomicData")

# Filter to samples of interest
my_samples <- sampleMetadata %>%
    filter(control %in% c("Case", "Study Control"),
           age >= 18,
           !is.na(sex))

# Retrieve relative abundance data
tse <- returnSamples(sample_data = my_samples[1:10, ],
                     data_type = "relative_abundance")
```

### Advanced Access with accessParquetData() and loadParquetData()

For more control over filtering and data loading:

```{r advanced_example, eval=FALSE}
# Connect to database
con <- accessParquetData(data_types = "relative_abundance")

# Apply filters and load
tse <- loadParquetData(con = con,
                       data_type = "relative_abundance",
                       filter_values = list(
                           clade_name_species = c("s__Streptococcus_mutans",
                                                  "s__Escherichia_coli")
                       ))
```

### Using Local Files

When working with downloaded parquet files:

```{r local_example, eval=FALSE}
# Point to local directory
con <- accessParquetData(file_paths = "path/to/parquet/files/",
                         data_types = "relative_abundance")

# Load data
tse <- loadParquetData(con = con,
                       data_type = "relative_abundance")
```

## Additional Resources

### Package Functions for Data Discovery

- `output_file_types()` - List available output file types and their properties
- `parquet_colinfo()` - Get column definitions for a specific data type
- `get_repo_info()` - List available Hugging Face repositories
- `get_ref_info()` - List available reference files
- `get_hf_parquet_urls()` - Get direct URLs to parquet files

### Vignettes

- **First 15 Minutes** - Quick start guide
- **Full Workflow** - Comprehensive data retrieval tutorial
- **Piecewise Workflow** - Advanced direct database access
- **Working with Large Parquet Files** - Strategies for large data types

### External Documentation

- [MetaPhlAn Documentation](https://github.com/biobakery/MetaPhlAn)
- [HUMAnN Documentation](https://github.com/biobakery/humann)
- [TreeSummarizedExperiment](https://bioconductor.org/packages/TreeSummarizedExperiment)

## Session Info

```{r session_info}
sessionInfo()
```
