---
title: "Common Issues"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Common Issues}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r general_setup, include=FALSE}
library(parkinsonsMetagenomicData)
library(dplyr)
library(DBI)
library(DT)
```

# Common Issues

## Large File Considerations

Handling the larger data types, like "genefamilies_stratified" and similar
files, works the exact same way as other data types but requires taking a few
factors into consideration. There are two main limiters:

1. Hugging Face rate limits
2. Result size

To avoid running into Hugging Face's rate limiting (HTTP 429 error), we want to
make sure that all filtering arguments are very selective and only executed on
sorted columns. Using "genefamilies_stratified" as an example, we can filter by
"uuid" and "gene_family_uniref", since those are the two parquet files
available. Filtering by "gene_family_species"
*while we are still accessing the parquet in a lazy manner* can throw an error
since a lot of the values in that column show up much more often in the dataset.
To filter these non-sorted columns, it is recommended to use alternate filters
that leverage the sorted columns (e.g., uuid) to create the initial subset and
TreeSummarizedExperiment, then apply the less selective filters to the now
locally available data.

Alternatively, we can download these large files. This will completely eliminate
the rate limits since we are no longer accessing them remotely.

```{r local_return_samples, eval=FALSE}
# Load sample info
sample_table <- sampleMetadata |>
    filter(study_name == "ZhangM_2023") |>
    select(where(~ !any(is.na(.x))))

# Load reference info
clade_name_ref <- load_ref("clade_name_ref")
feature_table <- clade_name_ref %>%
    filter(grepl("Faecalibacterium", clade_name_genus))

# Provide paths to local files
local_files <- c("/path/to/relative_abundance_clade_name_species.parquet",
                 "/path/to/relative_abundance_uuid.parquet")

experiment <- returnSamples(data_type = "relative_abundance",
                            sample_data = sample_table,
                            feature_data = feature_table,
                            repo = NULL,
                            local_files = local_files,
                            include_empty_samples = TRUE,
                            dry_run = FALSE)
```

These options do need to be balanced with your resources for storing and
handling both the raw and retrieved data, however. It may be helpful to place a
DuckDB database file in a location capable of large file storage, or run R
within a high-resource environment.

Below, a few example large file queries are executed. They may take a few
seconds to fully load the data.

```{r demo_large_queries}
# Establish connection to genefamilies_stratified files
con_gs <- accessParquetData(data_types = "genefamilies_stratified")

# Example query: we are looking for the below gene families within a particular
# study.
#
# UniRef90_T4BVE4 - present only in SPF mice
# UniRef90_A0A1B1SA57 - present only in WildR mice

# Pull study metadata
selected_samples <- sampleMetadata |>
    filter(study_name == "MazmanianS_DumitrescuDG") |>
    select(where(~ !any(is.na(.x))))

# Do as much as possible to narrow down sample IDs prior to filtering
wildr_ids <- selected_samples |>
    filter(uncurated_donor_microbiome_type == "WildR") |>
    dplyr::pull(uuid)

spf_ids <- selected_samples |>
    filter(uncurated_donor_microbiome_type == "SPF") |>
    dplyr::pull(uuid)

# Run the queries
spf_ex <- loadParquetData(con_gs, "genefamilies_stratified",
                          filter_values = list(gene_family_uniref = "UniRef90_T4BVE4",
                                               uuid = spf_ids))
spf_ex

wildr_ex <- loadParquetData(con_gs, "genefamilies_stratified",
                            filter_values = list(gene_family_uniref = "UniRef90_A0A1B1SA57",
                                                 uuid = wildr_ids))
wildr_ex

# The full sample ID list also works, but is a bit slower
all_ex <- loadParquetData(con_gs, "genefamilies_stratified",
                          filter_values = list(gene_family_uniref = c("UniRef90_T4BVE4", "UniRef90_A0A1B1SA57"),
                                               uuid = selected_samples$uuid))
all_ex
```

```{r session_info}
sessionInfo()
```
