---
title: "Available File Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Available File Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r chunk_set, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

# Available File Types

This is a brief overview of each type of file that `parkinsonsMetagenomicData`
is intended to work with. A summary table is available with `biobakery_files()`.
This guide examines the structure and contents of each file type, and provide
common use cases for each one.

## MetaPhlAn

For additional information, see the
[MetaPhlAn GitHub repo](https://github.com/biobakery/MetaPhlAn).

### Relative Abundance

This is the primary output for taxonomic profiling, showing the relative
abundance of each microbial taxon (from kingdom to species and strain level).
The base MetaPhlAn call to obtain this file is:

```{bash ex_relab_call}
$ metaphlan metagenome.fastq \
        --input_type fastq \
        -o profiled_metagenome.txt
```

Here are 15 rows of this file for one sample. It contains a header with
information on the database, exact `metaphlan` command, and number of reads
processed as well as the standard MetaPhlAn header row and column names.

```{text ex_relab_out}
#mpa_vJun23_CHOCOPhlAnSGB_202403
#/usr/local/bin/metaphlan --input_type bowtie2out --index latest --bowtie2db metaphlan --nproc 1 --unclassified_estimation -o metaphlan_unknown_list.tsv /dev/fd/63
#35820609 reads processed
#SampleID    Metaphlan_Analysis
#clade_name    NCBI_tax_id    relative_abundance    additional_species
UNCLASSIFIED    -1    14.32695    
k__Bacteria    2    85.33295166560956    
k__Archaea    2157    0.34009629850313816    
k__Bacteria|p__Firmicutes    2|1239    55.72767897077801    
k__Bacteria|p__Bacteroidota    2|976    27.98104021500391    
k__Bacteria|p__Proteobacteria    2|1224    1.2720648488663489    
k__Archaea|p__Euryarchaeota    2157|28890    0.34009629850313816    
k__Bacteria|p__Actinobacteria    2|201174    0.24810057959927395    
k__Bacteria|p__Bacteria_unclassified    2|    0.0951142178497579    
k__Bacteria|p__Bacteroidota|c__Bacteroidia|o__Bacteroidales|f__Bacteroidaceae|g__Bacteroides|s__Bacteroides_stercoris|t__SGB1830    2|976|200643|171549|815|816|46506|    0.19589142416994365    k__Bacteria|p__Bacteroidota|c__Bacteroidia|o__Bacteroidales|f__Bacteroidaceae|g__Bacteroides|s__Bacteroides_sp_NSJ_48
```

The columns contain the following information:

* clade_name: The taxonomic lineage of the detected microbial clade
* NCBI_tax_id: The NCBI Taxonomy identifier for the clade in clade_name
* relative_abundance: The proportion of the total microbial community
represented by the clade
* additional_species: Any other species represented by the same set of detected
markers

### Viral Clustering

This is the output file when performing Viral Sequence Clusters Analysis,
showing the coverage of viral genomes. The base MetaPhlAn call to obtain this
file is:

```{bash ex_vc_call}
$ metaphlan metagenome.fastq \
        --input_type fastq \
        --profile_vsc \
        --vsc_out profiled_viruses.txt \
        -o profiled_metagenome.txt
```

Here are 15 rows of this file for one sample. It contains a header with
information on the database, exact `metaphlan` command, and number of reads
processed as well as the standard MetaPhlAn header row and column names.

```{text ex_vc_out}
#mpa_vJun23_CHOCOPhlAnSGB_202403
#/usr/local/bin/metaphlan --input_type fastq --index latest --bowtie2db metaphlan --samout sam.bz2 --bowtie2out bowtie2.out --nproc 16 --profile_vsc --vsc_breadth 0.75 --vsc_out metaphlan_viruses_list.tsv -o metaphlan_bugs_list.tsv out.fastq
#SampleID    Metaphlan_Analysis
M-Group/Cluster    genomeName    len    breadth_of_coverage    depth_of_coverage_mean    depth_of_coverage_median    M-Group-Type [k|u]    First Genome in Cluster    Other Genomes
M697    VDB|0010-00F3-0-0001|M697-c3861-c6-c0    10022    1.0    2279.0755338255835    2374.0    uVSG    -    -
M441    VDB|0048-0014-0-0001|M441-c2353-c0-c2    8442    1.0    2474.0095948827293    1267.0    uVSG    -    -
M647    VDB|002C-0003-0-001D|M647-c2667-c1-c0    2631    1.0    117.01786393006462    111.0    uVSG    -    -
M282    VDB|001D-0115-0-0004|M282-c1770-c3-c58    1986    1.0    532.7703927492447    17.0    uVSG    -    -
M696    VDB|0047-0002-0-0004|M696-c2774-c1-c10    2219    0.9990986931050022    21.07577807848444    18.0    uVSG    -    -
M808    VDB|0010-0092-0-0000|M808-c3098-c0-c1    18422    0.9988057757029638    56.696902173913045    57.0    uVSG    -    -
M599    VDB|0003-0023-0-0000|M599-c2620-c0-c3    3025    0.9986776859504132    93.88282025819265    98.0    uVSG    -    -
M1075    VDB|0022-0016-0-0003|M1075-c3777-c1-c0    1606    0.9950186799501868    68.63329161451814    65.0    uVSG    -    -
M412    VDB|004F-0194-0-0003|M412-c2480-c2-c9    4124    0.9919980601357905    110.66071865069665    19.0    uVSG    -    -
M10    VDB|0010-004B-0-0002|M10-c1012-c0-c11    17451    0.9919202337974902    21.53056036972848    21.0    uVSG    -    -
M311    VDB|0026-0044-0-0000|M311-c1716-c0-c21    2349    0.9838229033631333    15.050194720900043    12.0    kVSG    NC_029014_Parabacteroides_phage_YZ2015b    NC_0290141_Parabacteroides_phage_YZ2015b
```

The columns contain the following information:

* M-Group/Cluster: internal group/cluster ID
* genomeName: name of the viral reference sequence as found in the database
* len: length of the detected viral sequence
* breadth_of_coverage: the proportion of the viral sequence detected
* depth_of_coverage_mean: the average number of reads covering each base in the
detected viral sequence
* depth_of_coverage_median: the median number of reads covering each base in the
detected viral sequence
* M-Group-Type [k|u]: is the group/cluster known (kVSG) or unknown (uVSG)
* First Genome in Cluster: for known groups/clusters, the representative genome
* Other Genomes: additional genomes belonging to the same group/cluster

### Marker Abundance

This file shows the abundance of individual marker genes within the sample,
expressed as normalized counts. The base MetaPhlAn call to obtain this file is:

```{bash ex_marker_ab_call}
$ metaphlan metagenome.fastq \
        --input_type fastq \
        -t marker_ab_table \
        -o marker_abundance.tsv \
```

Here are 15 rows of this file for one sample. It contains a header with
information on the database, exact `metaphlan` command, and number of reads
processed as well as the standard MetaPhlAn header row.

```{text ex_marker_ab_out}
#mpa_vJun23_CHOCOPhlAnSGB_202403
#/usr/local/bin/metaphlan --input_type bowtie2out --index latest --bowtie2db metaphlan -t marker_ab_table -o marker_abundance.tsv /dev/fd/63
#69691760 reads processed
#SampleID	Metaphlan_Analysis
UniClust90_BKJPBOJE01863|1__3|SGB29321	16.295999720557855
UniRef90_UPI0008F8481E|1__25|SGB29321	1.109008995129271
UniRef90_UPI0008F83D37|1__23|SGB29321	0.6048099752847368
UniRef90_UPI0008F879F5|1__23|SGB29321	0.3024049876423684
UniRef90_UPI0008F8EE49|1__23|SGB29321	1.2096199505694736
UniRef90_UPI0008F93A48|1__23|SGB29321	0.6048099752847368
UniRef90_UPI0008F89F0E|1__22|SGB29321	1.26709639083112
UniRef90_UPI0008F7EAE8|1__20|SGB29321	0.700078186137104
UniRef90_UPI0008F87225|3__21|SGB29321	1.4777467416595347
UniRef90_UPI0008F7E6FC|1__19|SGB29321	0.7388733708297673
UniRef90_UPI0008F84591|1__18|SGB29321	0.7822205139186101
```

Note that this file does not have an additional header row like other
MetPhlAn output file types. Due to this, the column names are not representative
of their contents. The columns contain the following information:

* SampleID: marker gene ID
* Metaphlan_Analysis: normalized marker coverage

### Marker Presence

This file is a list of markers present in the sample. The base MetaPhlAn call to
obtain this file is:

```{bash ex_marker_pres_call}
$ metaphlan metagenome.fastq \
        --input_type fastq \
        -t marker_pres_table \
        -o marker_presence.tsv \
```

Here are 15 rows of this file for one sample. It contains a header with
information on the database, exact `metaphlan` command, and number of reads
processed as well as the standard MetaPhlAn header row.

```{text ex_marker_pres_out}
#mpa_vJun23_CHOCOPhlAnSGB_202403
#/usr/local/bin/metaphlan --input_type bowtie2out --index latest --bowtie2db metaphlan -t marker_pres_table -o marker_presence.tsv /dev/fd/63
#34387180 reads processed
#SampleID	Metaphlan_Analysis
UniClust90_GMEPEMOD00192|1__6|SGB72336	1
UniClust90_HPNJJKHE00452|1__5|SGB72336	1
UniClust90_HBOCPNFF01148|1__5|SGB72336	1
UniRef90_A0A1I0IMR7|1__4|SGB4725	1
UniRef90_A0A1I0IRU8|1__7|SGB4725	1
UniRef90_A0A1I0AQT2|1__8|SGB4725	1
UniRef90_A0A1I0IRQ4|1__7|SGB4725	1
UniRef90_A0A3R6W3X6|1__6|SGB4725	1
UniRef90_A0A1I0JT88|1__4|SGB4725	1
UniRef90_A0A416TWF5|1__3|SGB4725	1
UniRef90_A0A1I0JQR0|1__9|SGB4725	1
```

Note that this file does not have an additional header row like other
MetPhlAn output file types. Due to this, the column names are not representative
of their contents. The columns contain the following information:

* SampleID: marker gene ID
* Metaphlan_Analysis: marker presence

### StrainPhlAn Markers

This file contains extra information on all of the assembled marker genes found
in a sample. They are used as input for future strain-level analysis. The base
MetPhlAn/StrainPhlAn calls to obtain this file are:

```{bash ex_strain_call}
$ metaphlan metagenome.fastq \
     --input_type fastq \
     --samout metagenome.sam.bz2 \
     -o profiled_metagenome.txt \

$ sample2markers.py \
     --input metagenome.sam.bz2 \
     --output_dir consensus_markers
```

Here are the first 15 rows of this file for one sample. It is in JSON format,
and contains the name of the database used to detect these markers.

```{text ex_strain_out}
{
  "database_name": "mpa_vJun23_CHOCOPhlAnSGB_202403",
  "consensus_markers": [
    {
      "marker": "UniRef90_A0A1I0JLZ1|1__3|SGB4725",
      "breath": 80.66666666666667,
      "avg_depth": 2.4962962962962965,
      "sequence": "--------------------------------------------------------------------GCTCAATGT-AATGAGATTATTGTGGCCAACGATCCGGTTTTAAAGGACGCCCTGCGCATCAGCCTGATGAAGATGACCGCCCCCCATGGCTGCAAGGTGAGGATCATTTCGGTGGCGGAGGCCGTCGGGTATCTGAACAGCCCTGAGGCAGAAAAGCGCAGGAT*CTGCTGTTGGTGGAGGGA-CCGCGGACGCGCTGCAGGTGGTGGAGGGGGTGAAGGGGATCACCCACTTGAACGCCGGGCTGATGAAAACCGGCGAGGGGAG*AAAATGGTCA**CCGACCCTGGC*TTTGCACCTGAGGATTTCGACAATTTTAAGAAAATCCTGGACCGGGGCGTGAG*GTGGAGTCCTATGTGACGCCCGATG-----------"
    },
    {
      "marker": "UniRef90_A0A416CGK3|6__9|SGB4993",
      "breath": 85.81818181818181,
      "avg_depth": 3.3181818181818183,
      "sequence": "--------------------------------------ATTCCATGATGAAGCATCCCCGTCCGACCCGTGCAGAGGCAACCGACGTTGCAAACGCTATCTACGATGGCACAACCGGTATCATGCTTTCCGGTGAGACAGCAAACGGCGATTACCCGCTGGAAGCTGTTCAGACCATGGTTCGTATCGCAGAGCGCGCAGAAAAGGATATCGACTATGTAGGCCGTCTGCAGAAGACTGGTGCAAGACTTCCTCAGGAGCAGGATACAACCACTTCTATCTGCCACGCAACCTGCACCGTTGCTACTGACCTGAATGCAGCAGCAATCATCCCGGTATCCATGTCTGGCTTTACTTCCGGCATGGTAGCACGTTTCAAACC*AACTGCCCGATCATCGCATGCACAACCAGCCGTCTGGTATGGCGTCAGATGAACCTGCAGTGGGGCGTAACTCCTCTGCTGATCGCAGAAGAGAACACTGCAGAGGATCTGTTCCGTGAAGCAGTTAAG---------------------------------------"
    },
```

The consensus marker details included are:

* marker: marker gene ID
* breath (breadth): proportion of the sequence covered
* avg_depth: average sequencing depth
* sequence: reconstructed consensus strain sequence

## HUMAnN

For additional information, see the
[HUMAnN GitHub repo](https://github.com/biobakery/humann).

### Genefamilies

This is a primary output for microbial functional profiling. It shows the
abundance of each gene family that is present in the microbial community. This
file type can be stratified or unstratified, meaning the abundance values are
subdivided or not based on the contributions of individual species. The
abundance values are reported as reads per kilobase (RPK), counts per million
(CPM), or relative abundance. The base HUMAnN calls to obtain these files are:

```{base ex_genefam_call}
# Base call; outputs values in RPK
$ humann \
        --input metagenome.fastq \
        --output output_dir

# Normalize to use CPM and relative abundance
$ humann_renorm_table \
        --input output_dir/out_genefamilies.tsv \
        --output output_dir/out_genefamilies_cpm.tsv \
        --units cpm

$ humann_renorm_table \
        --input output_dir/out_genefamilies.tsv \
        --output output_dir/out_genefamilies_relab.tsv \
        --units relab

# Split each table into one stratified table and one unstratified table
$ humann_split_stratified_table \
        --input output_dir/out_genefamilies.tsv \
        --output output_dir

$ humann_split_stratified_table \
        --input output_dir/out_genefamilies.tsv \
        --output output_dir

$ humann_split_stratified_table \
        --input output_dir/out_genefamilies.tsv \
        --output output_dir
```

Here are 15 example rows of the unstratified, unnormalized file for one sample.
It includes a basic HUMAnN header, which does not necessarily match the output
values. In this file the values reported are in fact in RPK, but files using CPM
or relative abundance will have the same header string.

```{text ex_genefam_unstrat_out}
# Gene Family	out_Abundance-RPKs
UNMAPPED	5154490.0000000000
UniRef90_A7AJP9	98090.8507772456
UniRef90_A0A174IWH6	61063.9749303064
UniRef90_A0A174PA28	51628.0623537519
UniRef90_A0A174Y3E0	43419.4009509238
UniRef90_R6HUL4	33652.7156806290
UniRef90_R5A5Y7	33591.5704773620
UniRef90_C4ZFG5	28701.2591291458
UniRef90_A7AJP8	28465.5157850624
UniRef90_A0A0T7A674	26531.0122765832
UniRef90_B0MYG3	26439.0665756726
UniRef90_V6DNH8	22177.9046993531
UniRef90_V8M0W9	21826.6862201161
UniRef90_A7V7B8	18829.5080249549
```

The columns contain the following information:

* Gene Family: The code of the detected gene family
* out_Abundance-RPKs: The abundance of the detected gene family in the total
microbial community, reported in RPK (reads per kilobase)

### Pathway Abundance

This file details the microbial pathways present in the community. This file
type can be stratified or unstratified, meaning the abundance values are
subdivided or not based on the contributions of individual species. The
abundance values are reported as reads per kilobase (RPK), counts per million
(CPM), or relative abundance. The base HUMAnN calls to obtain these files are:

```{base ex_pathab_call}
# Base call; outputs values in RPK
$ humann \
        --input metagenome.fastq \
        --output output_dir

# Normalize to use CPM and relative abundance
$ humann_renorm_table \
        --input output_dir/out_pathabundance.tsv \
        --output output_dir/out_pathabundance_cpm.tsv \
        --units cpm

$ humann_renorm_table \
        --input output_dir/out_pathabundance.tsv \
        --output output_dir/out_pathabundance_relab.tsv \
        --units relab

# Split each table into one stratified table and one unstratified table
$ humann_split_stratified_table \
        --input output_dir/out_pathabundance.tsv \
        --output output_dir

$ humann_split_stratified_table \
        --input output_dir/out_pathabundance.tsv \
        --output output_dir

$ humann_split_stratified_table \
        --input output_dir/out_pathabundance.tsv \
        --output output_dir
```

Here are 15 example rows of the unstratified, unnormalized file for one sample.
It includes a basic HUMAnN header. The header will again stay the same
regardless of actual output value type (RPK, CPM, or relative abundance).

```{text ex_pathab_unstrat_out}
# Pathway	out_Abundance
UNMAPPED	1963761.2729906905
UNINTEGRATED	3026993.4709948450
TRNA-CHARGING-PWY: tRNA charging	2697.9004105228
PWY-1042: glycolysis IV	2360.2965535212
PWY-6609: adenine and adenosine salvage III	2328.2392201336
PWY-7221: guanosine ribonucleotides de novo biosynthesis	2292.9280094236
PWY-6122: 5-aminoimidazole ribonucleotide biosynthesis II	2211.0246687537
PWY-6277: superpathway of 5-aminoimidazole ribonucleotide biosynthesis	2211.0246687537
PWY-5686: UMP biosynthesis I	2209.5938625315
PWY-7790: UMP biosynthesis II	2209.5938625315
PWY-7791: UMP biosynthesis III	2209.5938625315
PWY-5097: L-lysine biosynthesis VI	2174.1918575462
VALSYN-PWY: L-valine biosynthesis	2163.8066309442
PWY-6121: 5-aminoimidazole ribonucleotide biosynthesis I	2145.5730194284
```

The columns contain the following information:

* Pathway: The detected microbial pathway
* out_Abundance: The abundance of the detected microbial pathway within the
total microbial community, reported in RPK (reads per kilobase)

### Pathway Coverage

This file reports the proportion of genes within each microbial pathway that
were detected. A value of 1.0 means all genes in the pathway were found. This
file type can be stratified or unstratified, meaning the coverage values are
subdivided or not based on the contributions of individual species. The base
HUMAnN calls to obtain these files are:

```{base ex_pathcov_call}
# Base call
$ humann \
        --input metagenome.fastq \
        --output output_dir

# Split each table into one stratified table and one unstratified table
$ humann_split_stratified_table \
        --input output_dir/out_pathcoverage.tsv \
        --output output_dir
```

Here are 15 example rows of the unstratified file for one sample. It includes a
basic HUMAnN header.

```{text ex_pathcov_unstrat_out}
# Pathway	out_Coverage
UNMAPPED	1.0000000000
UNINTEGRATED	1.0000000000
PWY-5686: UMP biosynthesis I	1.0000000000
PWY-7790: UMP biosynthesis II	1.0000000000
PWY-7791: UMP biosynthesis III	1.0000000000
TRNA-CHARGING-PWY: tRNA charging	1.0000000000
DTDPRHAMSYN-PWY: dTDP-&beta;-L-rhamnose biosynthesis	1.0000000000
PWY-3841: folate transformations II (plants)	1.0000000000
PWY-6609: adenine and adenosine salvage III	1.0000000000
PWY-1042: glycolysis IV	1.0000000000
VALSYN-PWY: L-valine biosynthesis	1.0000000000
PEPTIDOGLYCANSYN-PWY: peptidoglycan biosynthesis I (meso-diaminopimelate containing)	1.0000000000
PWY-6700: queuosine biosynthesis I (de novo)	1.0000000000
ANAGLYCOLYSIS-PWY: glycolysis III (from glucose)	0.9017787803
```

The columns contain the following information:

* Pathway: The detected microbial pathway
* out_Coverage: The coverage of the detected microbial pathway within the
total microbial community, reported as a proportion

## Quality Control

### FastQC Metrics

For additional information, see the
[FastQC webpage](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/).

This file is simply the basic FastQC output text file. The call to obtain this
file is:

```{bash ex_fastqc_call}
$ fastqc seqfile
```

Here are the first 20 lines of the file for one sample. These show the Basic
Statistics module as well as the beginning of the Per Base Sequence Quality
module. Other standard modules can be found later in the file.

```{text ex_fastqc_out}
##FastQC	0.12.1
>>Basic Statistics	pass
#Measure	Value
Filename	out_sample.fastq
File type	Conventional base calls
Encoding	Sanger / Illumina 1.9
Total Sequences	50000
Total Bases	7.3 Mbp
Sequences flagged as poor quality	0
Sequence length	35-151
%GC	47
>>END_MODULE
>>Per base sequence quality	pass
#Base	Mean	Median	Lower Quartile	Upper Quartile	10th Percentile	90th Percentile
1	38.7797	40.0	40.0	40.0	40.0	40.0
2	38.10926	40.0	40.0	40.0	40.0	40.0
3	38.37914	40.0	40.0	40.0	40.0	40.0
4	38.20572	40.0	40.0	40.0	40.0	40.0
5	38.36706	40.0	40.0	40.0	40.0	40.0
6	38.77062	40.0	40.0	40.0	40.0	40.0
```

### KneadData Log

For additional information, see the
[KneadData GitHub repo](https://github.com/biobakery/kneaddata).

This file is the log file obtained when running KneadData. The base call to
obtain this file is:

```{bash ex_kneaddata_call}
$ kneaddata --unpaired seq.fastq \
    --reference-db $DATABASE \
    --output kneaddata_output
```

Here is one line from the log file. It is rather cluttered, but you can see the
information on input, surviving, and dropped reads that can be parsed from it.

```{text ex_kneaddata_out}
05/25/2025 02:03:00 AM - kneaddata.utilities - DEBUG: b"TrimmomaticSE: Started with arguments:\n -threads 16 -phred33 /shares/CIBIO-Storage/CM/scratch/users/kaelyn.long/cmd_nf/work/42/c00f840ea0a591c0c503b9b7889435/kneaddata_output/reformatted_identifiers3h38b_gw_decompressed_wi8x6cd8_out /shares/CIBIO-Storage/CM/scratch/users/kaelyn.long/cmd_nf/work/42/c00f840ea0a591c0c503b9b7889435/kneaddata_output/out_kneaddata.trimmed.fastq MINLEN:60 ILLUMINACLIP:/usr/local/lib/python3.9/site-packages/kneaddata/adapters/NexteraPE-PE.fa:2:30:10 SLIDINGWINDOW:4:20 MINLEN:74\nUsing PrefixPair: 'AGATGTGTATAAGAGACAG' and 'AGATGTGTATAAGAGACAG'\nUsing Long Clipping Sequence: 'GTCTCGTGGGCTCGGAGATGTGTATAAGAGACAG'\nUsing Long Clipping Sequence: 'TCGTCGGCAGCGTCAGATGTGTATAAGAGACAG'\nUsing Long Clipping Sequence: 'CTGTCTCTTATACACATCTCCGAGCCCACGAGAC'\nUsing Long Clipping Sequence: 'CTGTCTCTTATACACATCTGACGCTGCCGACGA'\nILLUMINACLIP: Using 1 prefix pairs, 4 forward/reverse sequences, 0 forward only sequences, 0 reverse only sequences\nInput Reads: 103418334 Surviving: 96226892 (93.05%) Dropped: 7191442 (6.95%)\nTrimmomaticSE: Completed successfully\n"
```

```{r sessionInfo, eval=TRUE}
sessionInfo()
```
