---
title: "First 15 Minutes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{First 15 Minutes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r general_setup, include=FALSE}
library(parkinsonsMetagenomicData)
library(TreeSummarizedExperiment)
library(dplyr)
```

# First 15 Minutes

This vignette is a fast path to retrieving microbiome data as a
`TreeSummarizedExperiment`.

## Vignette guide

| Vignette | Audience | Typical time | Use when |
|---|---|---:|---|
| `First 15 Minutes` | New users, biology-first workflows | 10-15 min | You want your first successful data retrieval quickly |
| `Full Workflow` | Intermediate/advanced users | 30-45 min | You want full control over sample, feature, and query setup |
| `Piecewise Workflow` | Advanced users | 30-60 min | You need direct control of `accessParquetData()` and `loadParquetData()` |
| `Common Issues` | Any user troubleshooting | 10-20 min | You hit rate limits, large-query issues, or performance bottlenecks |

## What you'll do

1. Pick samples from `sampleMetadata`.
2. Pick features from a reference table.
3. Retrieve data with `returnSamples()`.

## What you'll get

`returnSamples()` returns a `TreeSummarizedExperiment`:

- assay data (microbial measurements)
- sample metadata in `colData`
- feature metadata in `rowData`

## Quick start

To select a subset of available samples, start from the sample table:
```{r}
data("sampleMetadata", package = "parkinsonsMetagenomicData")
dim(sampleMetadata)
sampleMetadata[1:5, 1:10]
```

As an example, fetch MetaPhlAn4 relative abundance data for all 
Faecalibacterium species across all samples in the ZhangM_2023 study, 
directly from remote storage:

```{r quick_start, eval=interactive()}
# In any subset of samples there will be many rows of all NA,
# so the select function below is usually wanted.
sample_table <- sampleMetadata |>
    filter(study_name == "ZhangM_2023") |>
    select(where(~ !all(is.na(.x))))

# All terminal taxa from the Faecalibacterium genus.
feature_table <- load_ref("clade_name_ref") |>
    filter(grepl("Faecalibacterium", clade_name_genus)) |>
    filter(!is.na(clade_name_species)) |>
    filter(!is.na(clade_name_terminal))

tse <- returnSamples(
    data_type = "relative_abundance",
    sample_data = sample_table,
    feature_data = feature_table,
    repo = "waldronlab/metagenomics_mac",
    include_empty_samples = TRUE,
    dry_run = FALSE
)

tse
```

This just retrieved all Faecalibacterium species across all samples in the ZhangM_2023 study, 
downloading from a remote repo. You can select samples by phenotype instead of study, and select 
other types of features like metabolic functional modules inferred by HUMAnN. You can select a few
features across many samples, or many features across a few samples. If you want many features 
across many samples, you will need to download the data and query locally, which is covered in 
the `Piecewise Workflow` vignette.


## Inspect the result

```{r inspect_result, eval=interactive()}
assayNames(tse)
head(colData(tse)) # Rownames are sample unique identifiers (UUIDs)
head(rowData(tse)) #rownames are taxonomic lineages or functional modules
```

## If this fails

Important `?returnSamples` sections:
- `HTTP 429`, slow remote reads, or very large queries: download then use local parquet files with `local_files`.
- To inspect SQL before executing: set `dry_run = TRUE` and run `dplyr::show_query()`.

## Next step

- See `Full Workflow` for detailed configuration and discovery tooling.
- See `Piecewise Workflow` for manual control over connection and loading steps.
- See `Common Issues` for troubleshooting patterns and large-file strategy.

## Reference

- Metadata curation: https://github.com/ASAP-MAC/parkinsonsManualCuration
- Nextflow pipeline for processing metagenomic data: https://github.com/ASAP-MAC/metagenomicsNextflowMAC
- Datasets found in the literature and efforts to obtain them: https://github.com/ASAP-MAC/parkinsons_data_search


```{r session_info}
sessionInfo()
```
