---
title: "First 15 Minutes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{First 15 Minutes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r general_setup, include=FALSE}
library(parkinsonsMetagenomicData)
library(TreeSummarizedExperiment)
library(dplyr)
```

# First 15 Minutes

This vignette is a fast path to retrieving microbiome data as a
`TreeSummarizedExperiment`.

## Vignette guide

| Vignette | Audience | Typical time | Use when |
|---|---|---:|---|
| `Data Codebook` | All users | Reference | You need to understand variable definitions, data types, and column structures |
| `Available File Types` | All users | Reference | You want to understand the structure, columns, and examples of raw pipeline output files |
| `Full Data Preparation` | Advanced users/Developers | Reference | You want to understand how raw sequences were processed, transformed into Parquet files, and how metadata was curated |
| `First 15 Minutes` | New users, biology-first workflows | 10-15 min | You want your first successful data retrieval quickly |
| `Full Workflow` | Intermediate/advanced users | 30-45 min | You want full control over sample, feature, and query setup |
| `Piecewise Workflow` | Advanced users | 30-60 min | You need direct control of `accessParquetData()` and `loadParquetData()` |
| `Working with Large Parquet Files` | Users working with large data types | 10-20 min | You need to query genefamilies_stratified or other large files efficiently |

## What you'll do

1. Explore available data types and their descriptions.
2. Pick samples from `sampleMetadata`.
3. Pick features from a reference table.
4. Retrieve data with `returnSamples()`.

## What you'll get

`returnSamples()` returns a `TreeSummarizedExperiment`:

- assay data (microbial measurements)
- sample metadata in `colData`
- feature metadata in `rowData`

## Quick start

You can view metadata about all samples in `sampleMetaData`. Later in example 1, you'll see how to filter rows 
of this table to choose which samples you want to download data for. 
```{r load_sample_metadata}
data("sampleMetadata", package = "parkinsonsMetagenomicData")
dim(sampleMetadata)
sampleMetadata[1:5, 1:10]
```

## Exploring Available Data Types

Before retrieving data, you can explore what data types are available. Each data type represents different measurements or processing levels from various bioinformatics tools:

```{r explore_data_types}
available_data <- get_hf_parquet_urls()
```

```{r show_data_types, echo=FALSE}
library(DT)
get_hf_parquet_urls() |>
    select(data_type, tool, description, units_normalization) |>
    distinct() |>
    datatable(
        options = list(pageLength = 10),
        filter = 'top',
        caption = "Available data types in parkinsonsMetagenomicData"
    )
```

Use the search boxes above each column to filter by data type, tool, or keywords in the description. For example, search for "pathway" to find all pathway-related data types, or "CPM" to find normalized abundance measures.

## Quick start examples

### Example 1: Taxonomic relative abundance

Fetch MetaPhlAn4 relative abundance data for all Faecalibacterium species across all samples in the ZhangM_2023 study, directly from remote storage:

```{r quick_start}
# In any subset of samples there will be many rows of all NA,
# so the select function below is usually wanted.
sample_table <- sampleMetadata |>
    filter(study_name == "ZhangM_2023") |>
    select(where(~ !all(is.na(.x))))

# All terminal taxa from the Faecalibacterium genus.
feature_table <- load_ref("clade_name_ref") |>
    filter(grepl("Faecalibacterium", clade_name_genus)) |>
    filter(!is.na(clade_name_species)) |>
    filter(!is.na(clade_name_terminal))

tse <- returnSamples(
    data_type = "relative_abundance",
    sample_data = sample_table,
    feature_data = feature_table,
    repo = "waldronlab/metagenomics_mac",
    include_empty_samples = TRUE,
    dry_run = FALSE
)

tse
```

This just retrieved all Faecalibacterium species across all samples in the ZhangM_2023 study,
downloading from a remote repo.

### Example 2: Functional pathway data

You can also retrieve functional data like metabolic pathways. First, discover which reference tables are available:

```{r discover_refs}
# See what reference tables are available
ref_info <- get_ref_info()
```

```{r show_ref_info, echo=FALSE}
get_ref_info() |>
    datatable(
        options = list(pageLength = 10),
        filter = 'top',
        caption = "Available reference tables for feature discovery"
    )
```

Now load the pathway reference and browse for pathways of interest:

```{r browse_pathways}
# Load the pathway reference
pathway_ref <- load_ref("pathway_ref")

# Search for butyrate-related pathways (relevant to gut-brain health in PD)
butyrate_pathways <- pathway_ref |>
    filter(grepl("butanoate|butyrat", pathway, ignore.case = TRUE)) |>
    filter(!grepl("\\|", pathway)) |>  # Exclude stratified versions
    select(pathway) |>
    distinct()

butyrate_pathways
```

Now fetch the abundance of a specific butyrate biosynthesis pathway across all `r nrow(sampleMetadata)`
samples in parkinsonsMetagenomicData:

```{r pathway_example}
# Use all samples (no filtering by study)
# Select the acetyl-CoA fermentation to butanoate pathway
feature_table_pathway <- pathway_ref |>
    filter(pathway == "PWY-5676: acetyl-CoA fermentation to butanoate II") |>
    select(pathway)

tse_pathway <- returnSamples(
    data_type = "pathabundance_unstratified",
    sample_data = sampleMetadata,
    feature_data = feature_table_pathway,
    repo = "waldronlab/metagenomics_mac",
    include_empty_samples = TRUE,
    dry_run = FALSE
)

tse_pathway
```

You can select samples by phenotype instead of study, and select a few features across many samples, or many features across a few samples. If you want many features across many samples, you will need to download the data and query locally, which is covered in the `Piecewise Workflow` vignette.

## Inspect the result

Both examples return a `TreeSummarizedExperiment` object with assays, sample metadata, and feature metadata:

```{r inspect_result, eval=interactive()}
# Inspect the taxonomic data from Example 1
assayNames(tse)
head(colData(tse)) # Rownames are sample unique identifiers (UUIDs)
head(rowData(tse)) # Rownames are taxonomic lineages

# Inspect the pathway data from Example 2
assayNames(tse_pathway)
head(rowData(tse_pathway)) # Rownames are pathway IDs
```

## If this fails

If you `?returnSamples` fails:

- `HTTP 429`, slow internet access, or queries that will return a lot of data: download then use local parquet files with `local_files`.
- To inspect SQL before executing: set `dry_run = TRUE` and run `dplyr::show_query()`.

## Next step

- See `Full Workflow` for detailed configuration and discovery tooling.
- See `Piecewise Workflow` for manual control over connection and loading steps.
- See `Working with Large Parquet Files` for strategies on querying large data types like genefamilies_stratified.

## Reference

- Metadata curation: https://github.com/ASAP-MAC/parkinsonsManualCuration
- Nextflow pipeline for processing metagenomic data: https://github.com/ASAP-MAC/metagenomicsNextflowMAC
- Datasets found in the literature and efforts to obtain them: https://github.com/ASAP-MAC/parkinsons_data_search


```{r session_info}
sessionInfo()
```
