---
title: "Full Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r general_setup, include=FALSE}
library(parkinsonsMetagenomicData)
library(dplyr)
library(DBI)
library(DT)
```

# Parquet File Overview and Setup

## Hugging Face

Individual pipeline output files have been combined into parquet files
and hosted in the
[metagenomics_mac](https://huggingface.co/datasets/waldronlab/metagenomics_mac)
repo on Hugging Face. Smaller versions of those same files, comprising only 10
samples per file, are available in the Hugging Face repo
[metagenomics_mac_examples](https://huggingface.co/datasets/waldronlab/metagenomics_mac_examples).
These are publicly accessible, and are able to be easily read using DuckDB.

## Parquet Creation

To create the individual parquet files, one needs credentials to access the
`gs://metagenomics-mac` Google Bucket. Instructions on obtaining these can be
found in the
[Google Cloud Storage vignette](https://asap-mac.github.io/parkinsonsMetagenomicData/docs/vignettes/google_cloud_storage.html). These
credentials are then substituted into the scripts found in the repo
[parquet_generation](https://github.com/ASAP-MAC/parquet_generation). These
scripts perform a number of transformations that increase searchability before
storing the combined data for each data type in parquet files.

## DuckDB in R

DuckDB is very easy to use in R through the `duckdb` R package. The `DBI` and
`dplyr`/`dbplyr` packages combine with it to provide a streamlined way to work
with remote data by selectively querying it before bringing it into your R
session.

Relevant tools:

 - [DuckDB R Client](https://duckdb.org/docs/stable/clients/r.html)
 - [DBI](https://dbi.r-dbi.org/)
 - [dplyr](https://dplyr.tidyverse.org/)/[dbplyr](https://dbplyr.tidyverse.org/)


# Standard Workflow

In the standard workflow, we use the main wrapper function `returnSamples`. This
function takes tables with sample and feature information as well as a remote
repo name or a vector of paths to locally stored parquet files and retrieves the
relevant data as a TreeSummarizedExperiment.

Here is a brief overview of each argument to be provided to `returnSamples`,
please call `?returnSamples` for additional info.

 * **data_type**: the output file type of interest
 * **sample_data**: a table of sample metadata, used to specify which samples to
 retrieve data for
 * **feature_data**: a table of feature data, used to specify how to filter the raw
 data
 * **repo**: the identifier of a remote repo where the raw parquet files are stored
 * **local_files**: paths of locally stored parquet files, as an alternative to
 retrieval from a remote repo
 * **include_empty_samples**: sometimes none of the features specified in
 `feature_data` are found in one or more samples specified in `sample_data`.
 Should the samples still be included in the result with NA values for each
 feature (TRUE) or should they be omitted (FALSE)? Omitting these samples may
 increase response speed when accessing remote repos.
 * **dry_run**: a dry run returns a `tbl_duckdb_connection` object that contains all
 of the SQL code necessary to return the requested data. This SQL can be viewed
 by passing the object into `dplyr::show_query()`. This is useful for evaluating
 query efficiency prior to actually executing.

## File Selection

Both remote and local parquet files can be queried, though not at the same time.
`get_repo_info()` gives the names and URLs of the available repos, and
`get_hf_parquet_urls()` gives information on the files contained in those repos.
Those files can also be downloaded separately and then provided to the
`local_files` argument of `returnSamples()`. This may be desirable if internet
connection is unstable or limited, or if a particular query runs into the repo's
rate limits.

At this point, we also want to make a note of which `data_type` values we are
interested in. The `data_type` associated with each parquet file is listed in
the output of `get_hf_parquet_urls()`. For this example, we will be looking at
the 'relative_abundance' data type, which corresponds to the files
"relative_abundance_uuid.parquet" and
"relative_abundance_clade_name_species.parquet". The difference between these
two files is the column by which they are sorted:
"relative_abundance_uuid.parquet" is sorted by the 'uuid' column, while
"relative_abundance_clade_name_species.parquet" is sorted by the
'clade_name_species' column. Internal functions will choose which one to use,
we will simply provide "relative_abundance" as our chosen data type. If you
are downloading the files locally to avoid rate limits, you would download all
of the files associated with your chosen data type.

```{r demo_repo_info, eval=interactive()}
get_repo_info()
```
```{r show_repo_info, echo=FALSE}
get_repo_info() |>
    datatable(extensions = "Responsive")
```
```{r demo_parquet_urls, eval=interactive()}
get_hf_parquet_urls(repo_name = "waldronlab/metagenomics_mac")
```
```{r show_parquet_urls, echo=FALSE}
get_hf_parquet_urls(repo_name = "waldronlab/metagenomics_mac") |>
    suppressMessages() |>
    datatable(extensions = "Responsive")
```

## Sample Table

We can then examine the available samples and optionally select a subset to
query. This is done by browsing the `sampleMetadata` object. Here is an example
of how a set of data might be selected for a meta-analysis with some sampling
parameters. Alternatively, this step can be left off altogether. If the sample
table is not provided when calling `returnSamples()`, data for all samples will
simply be returned.

```{r pick_metadata}
## Load sampleMetadata
if (!exists("sampleMetadata", envir = environment())) {
    data("sampleMetadata", package = "parkinsonsMetagenomicData", envir = environment())
}

sample_table <- sampleMetadata |>
    filter(study_name == "ZhangM_2023") |>
    select(where(~ !any(is.na(.x))))
```
```{r show_metadata, echo=FALSE}
datatable(sample_table, extensions = "Responsive")
```

## Feature Table

For the feature table, we first have to find the reference file that goes with
the data type we are interested in. To do this, call `get_ref_info()`. Search
the 'general_data_type' column for your data type and identify the reference
file of interest. For a more granular view of what exactly will be in each file,
you can call `parquet_colinfo()` with your data type and see which columns will
be included. Here, we can see that "clade_name_ref" is associated with
"relative_abundance".

```{r demo_ref_info, eval=interactive()}
get_ref_info()
```
```{r show_ref_info, echo=FALSE}
get_ref_info() |>
    datatable(extensions = "Responsive")
```

And here we can see exactly which columns are in "clade_name_ref".

```{r demo_colinfo, eval=interactive()}
parquet_colinfo("relative_abundance")
```
```{r show_colinfo, echo=FALSE}
parquet_colinfo("relative_abundance") |>
    datatable(extensions = "Responsive")
```

Once we have found the correct reference file, we can filter it to only contain
features we are interested in. Here, we are interested in values for all
bacteria in the genus "Faecalibacterium". We load the reference file with
`load_ref()`, and use a basic `grepl()` filtering method.

```{r filter_feature_table}
clade_name_ref <- load_ref("clade_name_ref")
feature_table <- clade_name_ref %>%
    filter(grepl("Faecalibacterium", clade_name_genus))
```
```{r show_feature_table, echo=FALSE}
feature_table |>
    datatable(extensions = "Responsive")
```

As an extra consideration, MetaPhlAn relative abundance output includes
aggregate values for each taxonomic level. We can see above that the first row
returned has a value of `NA` for the "clade_name_species" column, and a few 
other rows have `NA` in the "clade_name_terminal" column. You may want
to remove these rows based on your analysis. To do so, we would simply re-filter
the file:

```{r re_filter_feature_table}
feature_table <- clade_name_ref %>%
    filter(grepl("Faecalibacterium", clade_name_genus)) %>%
    filter(!is.na(clade_name_species)) %>%
    filter(!is.na(clade_name_terminal))
```
```{r show_new_feature_table, echo=FALSE}
feature_table |>
    datatable(extensions = "Responsive")
```

## returnSamples()

We are now ready to retrieve our data. We simply pass our arguments into
`returnSamples()`. It may take a minute, depending on the amount of data
requested or your available resources.

```{r basic_return_samples}
experiment <- returnSamples(data_type = "relative_abundance",
                            sample_data = sample_table,
                            feature_data = feature_table,
                            repo = "waldronlab/metagenomics_mac",
                            local_files = NULL,
                            include_empty_samples = TRUE,
                            dry_run = FALSE)
experiment
```

If you are finding that this function is failing due to rate limiting (HTTP 429
error), hanging, or simply taking longer than you would like, you can download
the files associated with the data type you are interested in and supply their
paths to the "local_files" argument in lieu of specifying the "repo" argument.
It would look something like this:

```{r local_return_samples, eval=interactive()}
local_files <- c("/path/to/relative_abundance_clade_name_species.parquet",
                "/path/to/relative_abundance_uuid.parquet")

experiment <- returnSamples(data_type = "relative_abundance",
                            sample_data = sample_table,
                            feature_data = feature_table,
                            repo = NULL,
                            local_files = local_files,
                            include_empty_samples = TRUE,
                            dry_run = FALSE)
```

Finally, you can check exactly which query is being called on the raw parquet
data by passing "TRUE" to the "dry_run" argument. This returns a
`tbl_duckdb_connection` object that can be passed to `dplyr::show_query()`. To
demonstrate:

```{r query_return_samples}
query_only <- returnSamples(data_type = "relative_abundance",
                            sample_data = sample_table,
                            feature_data = feature_table,
                            repo = "waldronlab/metagenomics_mac",
                            local_files = NULL,
                            include_empty_samples = FALSE,
                            dry_run = TRUE)
dplyr::show_query(query_only)
```

```{r session_info}
sessionInfo()
```
