---
title: "Full Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r general_setup, include=FALSE}
library(parkinsonsMetagenomicData)
library(dplyr)
library(DBI)
library(DT)
```

# Parquet File Overview and Setup

## Hugging Face

Individual pipeline output files have been combined into parquet files
and hosted in the
[metagenomics_mac](https://huggingface.co/datasets/waldronlab/metagenomics_mac)
repo on Hugging Face. Smaller versions of those same files, comprising only 10
samples per file, are available in the Hugging Face repo
[metagenomics_mac_examples](https://huggingface.co/datasets/waldronlab/metagenomics_mac_examples).
These are publicly accessible, and are able to be easily read using DuckDB.

## DuckDB in R

DuckDB is very easy to use in R through the `duckdb` R package. The `DBI` and
`dplyr`/`dbplyr` packages combine with it to provide a streamlined way to work
with remote data by selectively querying it before bringing it into your R
session.

Relevant tools:

 - [DuckDB R Client](https://duckdb.org/docs/stable/clients/r.html)
 - [DBI](https://dbi.r-dbi.org/)
 - [dplyr](https://dplyr.tidyverse.org/)/[dbplyr](https://dbplyr.tidyverse.org/)


# Standard Workflow

In the standard workflow, we use the main wrapper function `returnSamples`. This
function takes tables with sample and feature information as well as a remote
repo name or a vector of paths to locally stored parquet files and retrieves the
relevant data as a TreeSummarizedExperiment.

Here is a brief overview of each argument to be provided to `returnSamples`,
please call `?returnSamples` for additional info.

 * **data_type**: the output file type of interest
 * **sample_data**: a table of sample metadata, used to specify which samples to
 retrieve data for
 * **feature_data**: a table of feature data, used to specify how to filter the raw
 data
 * **repo**: the identifier of a remote repo where the raw parquet files are stored
 * **local_files**: paths of locally stored parquet files, as an alternative to
 retrieval from a remote repo
 * **include_empty_samples**: sometimes none of the features specified in
 `feature_data` are found in one or more samples specified in `sample_data`.
 Should the samples still be included in the result with NA values for each
 feature (TRUE) or should they be omitted (FALSE)? Omitting these samples may
 increase response speed when accessing remote repos.
 * **dry_run**: a dry run returns a `tbl_duckdb_connection` object that contains all
 of the SQL code necessary to return the requested data. This SQL can be viewed
 by passing the object into `dplyr::show_query()`. This is useful for evaluating
 query efficiency prior to actually executing.

## File Selection

Both remote and local parquet files can be queried, though not at the same time.
`get_repo_info()` gives the names and URLs of the available repos, and
`get_hf_parquet_urls()` gives information on the files contained in those repos.
Those files can also be downloaded separately and then provided to the
`local_files` argument of `returnSamples()`. This may be desirable if internet
connection is unstable or limited, or if a particular query runs into the repo's
rate limits.

At this point, we also want to make a note of which `data_type` values we are
interested in. The `data_type` associated with each parquet file is listed in
the output of `get_hf_parquet_urls()`. For this example, we will be looking at
the 'relative_abundance' data type, which corresponds to the files
"relative_abundance_uuid.parquet" and
"relative_abundance_clade_name_species.parquet". The difference between these
two files is the column by which they are sorted:
"relative_abundance_uuid.parquet" is sorted by the 'uuid' column, while
"relative_abundance_clade_name_species.parquet" is sorted by the
'clade_name_species' column. Internal functions will choose which one to use,
we will simply provide "relative_abundance" as our chosen data type. If you
are downloading the files locally to avoid rate limits, you would download all
of the files associated with your chosen data type.

```{r demo_repo_info, eval=interactive()}
get_repo_info()
```
```{r show_repo_info, echo=FALSE}
get_repo_info() |>
    datatable(extensions = "Responsive")
```
```{r demo_parquet_urls, eval=interactive()}
get_hf_parquet_urls(repo_name = "waldronlab/metagenomics_mac")
```
```{r show_parquet_urls, echo=FALSE}
get_hf_parquet_urls(repo_name = "waldronlab/metagenomics_mac") |>
    suppressMessages() |>
    datatable(extensions = "Responsive")
```

### Exploring Data Types

For easier browsing, we can view just the unique data types with their descriptions:

```{r explore_data_types, eval=interactive()}
available_data <- get_hf_parquet_urls(repo_name = "waldronlab/metagenomics_mac")
```

```{r show_data_types, echo=FALSE}
get_hf_parquet_urls(repo_name = "waldronlab/metagenomics_mac") |>
    suppressMessages() |>
    select(data_type, tool, description, units_normalization) |>
    distinct() |>
    datatable(
        options = list(pageLength = 10),
        filter = 'top',
        caption = "Available data types with descriptions",
        extensions = "Responsive"
    )
```

Use the search boxes to filter by keywords. For example:
- Search "pathway" to find pathway abundance and coverage data
- Search "stratified" to find species-level breakdowns
- Search "viral" to find viral cluster data
- Search "genefamilies" to find gene family data

## Sample Table

We can then examine the available samples and optionally select a subset to
query. This is done by browsing the `sampleMetadata` object. Here is an example
of how a set of data might be selected for a meta-analysis with some sampling
parameters. Alternatively, this step can be left off altogether. If the sample
table is not provided when calling `returnSamples()`, data for all samples will
be returned.

```{r pick_metadata}
data("sampleMetadata", package = "parkinsonsMetagenomicData")
sample_table <- sampleMetadata |>
    filter(study_name == "ZhangM_2023") |>
    select(where(~ !any(is.na(.x))))
```
```{r show_metadata, echo=FALSE}
datatable(sample_table, extensions = "Responsive")
```

## Feature Table

For the feature table, we first have to find the reference file that goes with
the data type we are interested in. To do this, call `get_ref_info()`. Search
the 'general_data_type' column for your data type and identify the reference
file of interest. For a more granular view of what is in each file,
you can call `parquet_colinfo()` with your data type and see which columns will
be included. Here, we can see that "clade_name_ref" is associated with
"relative_abundance".

```{r demo_ref_info, eval=interactive()}
get_ref_info()
```
```{r show_ref_info, echo=FALSE}
get_ref_info() |>
    datatable()
```

### Understanding Reference Table Structure

To see exactly which columns are available in the reference table and understand how you can filter your features, use `parquet_colinfo()`. This shows all columns in both the data files and their associated reference table:

```{r demo_colinfo, eval=interactive()}
parquet_colinfo("relative_abundance")
```
```{r show_colinfo, echo=FALSE}
parquet_colinfo("relative_abundance") |>
    datatable(extensions = "Responsive")
```

This information is useful for:

- **Filtering features**: Identify which columns you can use to filter your feature table (e.g., `clade_name_genus`, `clade_name_species`, `clade_name_terminal`)
- **Understanding data structure**: See what metadata is available for each feature
- **Verifying your reference**: Confirm you're using the correct reference table for your data type

Once we have found the correct reference file and understand its structure, we can filter it to only contain features we are interested in. Here, we are interested in values for all
bacteria in the genus "Faecalibacterium". We load the reference file with
`load_ref()`, and use a basic `grepl()` filtering method.

```{r filter_feature_table}
clade_name_ref <- load_ref("clade_name_ref")
feature_table <- clade_name_ref %>%
    filter(grepl("Faecalibacterium", clade_name_genus))
```
```{r show_feature_table, echo=FALSE}
feature_table |>
    datatable(extensions = "Responsive")
```

As an extra consideration, MetaPhlAn relative abundance output includes
aggregate values for each taxonomic level. We can see above that the first row
returned has a value of `NA` for the "clade_name_species" column, and a few 
other rows have `NA` in the "clade_name_terminal" column. You may want
to remove these rows based on your analysis. To do so, we would simply re-filter
the file:

```{r re_filter_feature_table}
feature_table <- clade_name_ref %>%
    filter(grepl("Faecalibacterium", clade_name_genus)) %>%
    filter(!is.na(clade_name_species)) %>%
    filter(!is.na(clade_name_terminal))
```
```{r show_new_feature_table, echo=FALSE}
feature_table |>
    datatable(extensions = "Responsive")
```

## returnSamples()

We are now ready to retrieve our data. We simply pass our arguments into
`returnSamples()`. It may take a minute, depending on the amount of data
requested or your available resources.

```{r basic_return_samples}
experiment <- returnSamples(data_type = "relative_abundance",
                            sample_data = sample_table,
                            feature_data = feature_table,
                            repo = "waldronlab/metagenomics_mac",
                            local_files = NULL,
                            include_empty_samples = TRUE,
                            dry_run = FALSE)
experiment
```

If you are finding that this function is failing due to rate limiting (HTTP 429
error), hanging, or simply taking longer than you would like, you can download
the files associated with the data type you are interested in and supply their
paths to the "local_files" argument in lieu of specifying the "repo" argument.
Recall that you can use `get_hf_parquet_urls(repo_name = "waldronlab/metagenomics_mac")`
to find URLs to download full files.

Here's a working example using the small example parquet files included in the package:

```{r local_return_samples}
# Locate the example parquet files included in the package
local_files <- c(
    file.path(system.file("extdata", package = "parkinsonsMetagenomicData"),
              "pathcoverage_unstratified_pathway.parquet"),
    file.path(system.file("extdata", package = "parkinsonsMetagenomicData"),
              "pathcoverage_unstratified_uuid.parquet")
)

# Load the pathway reference for pathcoverage data
pathway_ref_local <- load_ref("pathway_ref",
                               file_path = file.path(system.file("extdata",
                                                     package = "parkinsonsMetagenomicData"),
                                                     "pathway_ref.parquet"))

# Select a few pathways
feature_table_local <- pathway_ref_local |>
    head(5) |>
    select(pathway)

# The example files contain data for 10 samples
# For this example, we can query all samples by omitting sample_data,
# or specify a subset using UUIDs
sample_uuids <- c("8793b1dc-3ba1-4591-82b8-4297adcfa1d7",
                  "cc1f30a0-45d9-41b1-b592-7d0892919ee7",
                  "fb7e8210-002a-4554-b265-873c4003e25f")
sample_table_local <- data.frame(uuid = sample_uuids)

# Query the local files
experiment_local <- returnSamples(
    data_type = "pathcoverage_unstratified",
    sample_data = sample_table_local,
    feature_data = feature_table_local,
    repo = NULL,
    local_files = local_files,
    include_empty_samples = TRUE,
    dry_run = FALSE
)

experiment_local
```

This demonstrates the complete local file workflow using the small example files included with the package. The example files contain 10 samples and can be queried just like full-sized parquet files. For larger datasets, you would download the full parquet files from Hugging Face and provide their paths in the same way.

Finally, you can check exactly which query is being called on the raw parquet
data by passing "TRUE" to the "dry_run" argument. This returns a
`tbl_duckdb_connection` object that can be passed to `dplyr::show_query()`. To
demonstrate:

```{r query_return_samples}
query_only <- returnSamples(data_type = "relative_abundance",
                            sample_data = sample_table,
                            feature_data = feature_table,
                            repo = "waldronlab/metagenomics_mac",
                            local_files = NULL,
                            include_empty_samples = FALSE,
                            dry_run = TRUE)
dplyr::show_query(query_only)
```

# Additional Data Type Examples

The following examples demonstrate retrieving different types of metagenomic data beyond taxonomic relative abundance. Each showcases unique features of the package and different bioinformatics tools.

## Example 2: Pathway Abundance (Stratified vs Unstratified)

HUMAnN pathway data comes in two forms:
- **Unstratified**: Total pathway abundance across all contributing organisms
- **Stratified**: Pathway abundance broken down by individual species

This example shows how to retrieve unstratified pathway abundance for butyrate biosynthesis pathways, which are relevant to gut-brain health in Parkinson's Disease.

### Discover Reference Tables

```{r pathway_ref_info, eval=interactive()}
get_ref_info()
```

```{r show_pathway_ref_info, echo=FALSE}
get_ref_info() |>
    filter(grepl("pathway", ref_file, ignore.case = TRUE)) |>
    datatable(extensions = "Responsive",
              caption = "Pathway reference tables")
```

### Browse Available Pathways

```{r browse_pathways}
pathway_ref <- load_ref("pathway_ref")

# Search for butyrate-related pathways
butyrate_pathways <- pathway_ref |>
    filter(grepl("butanoate|butyrat", pathway, ignore.case = TRUE)) |>
    filter(!grepl("\\|", pathway)) |>  # Exclude stratified versions (contains "|")
    select(pathway) |>
    distinct()
```

```{r show_butyrate_pathways, echo=FALSE}
butyrate_pathways |>
    datatable(extensions = "Responsive",
              caption = "Butyrate-related pathways (unstratified)")
```

### Retrieve Unstratified Pathway Data

```{r pathway_unstratified_example}
# Select a single pathway
feature_table_pathway <- pathway_ref |>
    filter(pathway == "PWY-5676: acetyl-CoA fermentation to butanoate II") |>
    select(pathway)

# Use samples from a single study
sample_table_pathway <- sampleMetadata |>
    filter(study_name == "ZhangM_2023") |>
    select(where(~ !any(is.na(.x))))

# Retrieve unstratified pathway abundance
tse_pathway <- returnSamples(
    data_type = "pathabundance_unstratified",
    sample_data = sample_table_pathway,
    feature_data = feature_table_pathway,
    repo = "waldronlab/metagenomics_mac",
    include_empty_samples = TRUE,
    dry_run = FALSE
)

tse_pathway
```

### Understanding Stratified Data

If you wanted to see WHICH bacterial species contribute to this pathway's abundance, you would use `data_type = "pathabundance_stratified"` instead. Stratified pathways have the format `PATHWAY|SPECIES`, for example:

```
PWY-5676: acetyl-CoA fermentation to butanoate II|g__Faecalibacterium.s__Faecalibacterium_prausnitzii
```

```{r pathway_stratified_example, eval=interactive()}
# Example: Get stratified version (species-level contributions)
# Note: This may return more rows as each pathway-species pair is a feature
feature_table_stratified <- pathway_ref |>
    filter(grepl("^PWY-5676:", pathway)) |>
    filter(grepl("Faecalibacterium", pathway)) |>  # Just Faecalibacterium contributions
    select(pathway)

tse_pathway_stratified <- returnSamples(
    data_type = "pathabundance_stratified",
    sample_data = sample_table_pathway,
    feature_data = feature_table_stratified,
    repo = "waldronlab/metagenomics_mac",
    include_empty_samples = TRUE,
    dry_run = FALSE
)
```

## Example 3: Viral Clusters

MetaPhlAn can identify viral sequences in metagenomic samples. This example retrieves viral cluster data.

### Discover Viral Reference

```{r viral_ref_info}
genome_name_ref <- load_ref("genome_name_ref")

# Browse available viral genomes
viral_genomes <- genome_name_ref |>
    select(genome_name) |>
    distinct() |>
    head(20)  # Show first 20 for brevity
```

```{r show_viral_genomes, echo=FALSE}
viral_genomes |>
    datatable(extensions = "Responsive",
              caption = "Example viral genome references (first 20)")
```

### Retrieve Viral Cluster Data

```{r viral_example}
# Select a few viral genomes to query
feature_table_viral <- genome_name_ref |>
    head(5) |>  # Select first 5 viral genomes
    select(genome_name)

# Use the same sample set
tse_viral <- returnSamples(
    data_type = "viral_clusters",
    sample_data = sample_table,
    feature_data = feature_table_viral,
    repo = "waldronlab/metagenomics_mac",
    include_empty_samples = TRUE,
    dry_run = FALSE
)

tse_viral
```

## Example 4: Gene Family Abundance

HUMAnN quantifies gene families (groups of homologous genes) using the UniRef database. This example shows gene family abundance data.

**Note**: Gene family queries can be large. For demonstration, we'll query a small number of gene families.

### Discover Gene Family Reference

```{r gene_ref_info}
gene_family_ref <- load_ref("gene_family_ref")

# Browse the structure
gene_family_ref |>
    select(gene_family) |>
    head(20)
```

```{r show_gene_families, echo=FALSE}
gene_family_ref |>
    head(20) |>
    datatable(extensions = "Responsive",
              caption = "Example gene families (first 20)")
```

### Retrieve Gene Family Data

```{r gene_family_example}
# Select specific gene families (using a small set for performance)
feature_table_genes <- gene_family_ref |>
    head(10) |>  # Select first 10 gene families
    select(gene_family)

# Use fewer samples for gene family queries (they can be large)
sample_table_small <- sampleMetadata |>
    filter(study_name == "ZhangM_2023") |>
    head(5) |>  # Use only 5 samples
    select(where(~ !any(is.na(.x))))

tse_genes <- returnSamples(
    data_type = "genefamilies_unstratified",
    sample_data = sample_table_small,
    feature_data = feature_table_genes,
    repo = "waldronlab/metagenomics_mac",
    include_empty_samples = TRUE,
    dry_run = FALSE
)

tse_genes
```

**Large Query Considerations**: Gene family stratified data (`genefamilies_stratified`) can be very large. For large queries, consider:
1. Download parquet files locally and use them via the `local_files` argument
2. Use very selective filters (specific gene families + specific samples)
3. Query in batches
4. See the `Common Issues` vignette for more strategies

# Comparing Data Types

Here's a summary of what we retrieved in these examples:

| Example | Data Type | Tool | Features | What It Measures |
|---------|-----------|------|----------|------------------|
| 1 | relative_abundance | MetaPhlAn | Bacterial taxa | Which bacteria are present and their relative abundance |
| 2 | pathabundance_unstratified | HUMAnN | Metabolic pathways | Total abundance of metabolic pathways |
| 3 | viral_clusters | MetaPhlAn | Viral genomes | Presence/abundance of viral sequences |
| 4 | genefamilies_unstratified | HUMAnN | Gene families | Abundance of functional gene groups |

All return `TreeSummarizedExperiment` objects with the same structure, making it easy to apply consistent analysis workflows across different data types.

```{r session_info}
sessionInfo()
```
