---
title: "Working With Data Hosted on Hugging Face"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working With Data Hosted on Hugging Face}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

```{r knitr_setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r general_setup, include=FALSE}
library(parkinsonsMetagenomicData)
library(dplyr)
library(DT)
```

# Hugging Face Overview and Setup

## Hugging Face

Individual pipeline output files have been combined into parquet files
and hosted in the
[metagenomics_mac](https://huggingface.co/datasets/waldronlab/metagenomics_mac)
repo on Hugging Face. *Note: some output file types are still pending.*
These are publicly accessible, and are able to be easily read using DuckDB.

## Parquet Creation

To create the individual parquet files, one needs credentials to access the
`gs://metagenomics-mac` Google Bucket. Instructions on obtaining these can be
found in the
[Google Cloud Storage vignette](vignettes/google_cloud_storage.Rmd). These
credentials are then substituted into the scripts found in the repo
[parquet_generation](https://github.com/ASAP-MAC/parquet_generation). These
scripts perform a number of transformations that increase searchability before
storing the combined data for each data type in parquet files.

## DuckDB in R

DuckDB is very easy to use in R through the `duckdb` R package. The `DBI` and
`dplyr`/`dbplyr` packages combine with it to provide a streamlined way to work
with remote data by selectively querying it before bringing it into your R
session.

Relevant tools:

 - [DuckDB R Client](https://duckdb.org/docs/stable/clients/r.html)
 - [DBI](https://dbi.r-dbi.org/)
 - [dplyr](https://dplyr.tidyverse.org/)/[dbplyr](https://dbplyr.tidyverse.org/)

# Standard Workflow

In the standard workflow, we have two main wrapper functions `accessParquetData`
and `loadParquetData`. `accessParquetData` sets up a DuckDB connection to the
parquet files hosted in Hugging Face. `loadParquetData` then takes an argument
for data type, and optionally arguments for specific sample UUIDs or a custom
filtering function and loads the requested data into R as a Tree Summarized
Experiment.

## Selecting Samples

First we select samples by browsing the `sampleMetadata` object.

```{r pick_metadata}
selected_samples <- sampleMetadata |>
    filter(study_name == "ZhangM_2023") |>
    select(where(~ !any(is.na(.x))))
```

```{r show_metadata, echo=FALSE}
datatable(selected_samples, extensions = "Responsive")
```

The `uuid` column contains the UUIDs we will need for `loadParquetData`.

```{r pull_uuids}
selected_uuids <- selected_samples$uuid
selected_uuids
```

## Hugging Face and DuckDB Connection

We then run `accessParquetData`, which creates the DuckDB connection and sets
up DuckDB "VIEW" objects for each available file type. It returns a DuckDB
connection object.

```{r access_wrapper}
con <- accessParquetData()
```

## Loading into R

We then provide that DuckDB connection object, our choice of data type/view, and
our list of UUIDs to `loadParquetData` and receive a Tree Summarized Experiment
object.

```{r load_wrapper}
basic_se <- loadParquetData(con,
                            data_type = "pathcoverage_unstratified",
                            uuids = selected_uuids)
basic_se
```

## Optional View Customization

Additionally, it is possible to use `dplyr` to preview the data view of interest
and perform more advanced functions on the data prior to using `loadParquetData`
or the `collect()` function to load it into R. Here we use the `tbl()` function
to access the data view of choice, and pipe it into a `filter()` call to select
only the pathways related to UMP biosynthesis. We can then save these calls to a
variable prior to loading them into R, and supply this to `loadParquetData`
either instead of or alongside our UUID list.

```{r custom_view}
custom_filter <- tbl(con, "pathcoverage_unstratified") |>
    filter(grepl("UMP biosynthesis", `# Pathway`))
custom_filter
```

```{r load_custom}
custom_se <- loadParquetData(con,
                            data_type = "pathcoverage_unstratified",
                            uuids = selected_uuids,
                            custom_view = custom_filter)
custom_se
```
