---
title: "parkinsonsMetagenomicData"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{parkinsonsMetagenomicData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages({
    library(parkinsonsMetagenomicData)
    library(dplyr)
    library(DT)
})
```

# Sample Metadata

Metadata for all samples present within `parkinsonsMetagenomicData` is available
through the included `data.frame` `sampleMetadata`. Both curated and uncurated
features are included in this `data.frame`, with uncurated features being
prefixed by "uncurated_". Curated features include the following at this time:

 * curation_id
 * study_name
 * sample_id
 * subject_id
 * target_condition
 * target_condition_ontology_term_id
 * control
 * control_ontology_term_id
 * age
 * age_group
 * age_group_ontology_term_id
 * age_unit
 * age_unit_ontology_term_id
 * sex
 * sex_ontology_term_id
 * disease
 * disease_ontology_term_id
 * curator
 * BioProject
 * BioSample
 * NCBI_accession
 * uuid

```{r}
sampleMetadata |>
    filter(study_name == "BedarfJR_2017") |>
    select(where(~ !any(is.na(.x)))) |>
    slice(1:10) |>
    select(1:10) |>
    datatable(options = list(dom = "t"), extensions = "Responsive")
```

# MetaPhlAn Results

The UUID(s) of any sample(s) of interest may be then used to access MetaPhlAn
output files.

## Google Bucket Setup

Output files are contained in the private Google Cloud Bucket
`gs://metagenomics-mac` at the moment, the user must authenticate with a service
account keyfile in the following way:

```{r, eval=FALSE}
googleCloudStorageR::gcs_auth("full/path/to/keyfile.json")
```

To obtain this keyfile, the owner of a service account affiliated with the
Google Cloud Project containing `gs://metagenomics-mac` must create it
according to the process detailed in the Google Cloud Guide
"[Create and delete service account keys](https://cloud.google.com/iam/docs/keys-create-delete)".

Additionally, the default bucket for `GoogleCloudStorageR` must be set to the
Google Bucket name, in this case `metagenomics-mac`. This is done automatically
upon loading the package, but if an error occurs can be manually done through
the following:

```{r, eval=FALSE}
googleCloudStorageR::gcs_global_bucket("metagenomics-mac")
```

## Data Retrieval

All objects stored in `gs://metagenomics-mac` can then be viewed with the
following. This operation can take several minutes due to the number of objects
stored in `gs://metagenomics-mac`.

```{r, eval=FALSE}
googleCloudStorageR::gcs_list_objects()
```

Alternatively, `listMetagenomicData` will provide a table of objects that are compatible
with the `getMetagenomicData` streamlined downloading and caching function.

```{r}
file_tbl <- listMetagenomicData()
```

```{r, echo=FALSE}
datatable(file_tbl[1:10,], options = list(dom = "t"), extensions = "Responsive")
```

To access MetaPhlAn output files, use `getMetagenomicData`. This function takes
UUID and data type arguments, downloads the corresponding MetaPhlAn output
files, and stores them in a local cache. If the same files are requested again
through this function, they will not be re-downloaded unless explicitly
specified, in order to reduce excessive downloads. `getMetagenomicData` returns
a list of SummarizedExperiment objects if `load = TRUE`, or a tibble of cached
file paths and cache IDs for each requested file if `load = FALSE`. A single
metaphlan list file can then be parsed into a SummarizedExperiment object
separately with `parse_metaphlan_list`. Additionally, sample metadata will be
included within the SummarizedExperiment objects as colData if `load = TRUE`, 
but will have to be added separately if `load = FALSE`.

```{r}
cache_tbl <- getMetagenomicData(uuids = "004c5d07-ec87-40fe-9a72-6b23d6ec584e",
                                data_types = "all",
                                load = FALSE)
```

```{r, echo=FALSE}
datatable(cache_tbl, options = list(dom = "t"), extensions = "Responsive")
```

## Data Handling

### Stepwise Experiment Setup

With the cache information obtained above, we can manually load in individual
files and format them as SummarizedExperiment objects complete with sample
metadata. `parse_metaphlan_list` completes the first step for files with
'data_type' equal to "bugs", "viruses", or "unknown".

```{r}
parsed_bugs_list <- parse_metaphlan_list(sample_id = cache_tbl$UUID[1],
                                         file_path = cache_tbl$cache_path[1],
                                         data_type = cache_tbl$data_type[1])
parsed_bugs_list
```

The corresponding metadata from sampleMetadata is then added as colData with
the function `add_metadata`.

```{r}
bugs_with_metadata <- add_metadata(sample_ids = colnames(parsed_bugs_list),
                                   id_col = "uuid",
                                   experiment = parsed_bugs_list)
bugs_with_metadata
```

### Automatic Experiment Setup

The above steps can also be accomplished automatically with `getMetagenomicData`
with the parameter `load = TRUE`.

```{r}
sample_experiments <- getMetagenomicData(uuids = c("0001a4de-a907-46bd-8523-d7bfc1cdb544",
                                                  "004c5d07-ec87-40fe-9a72-6b23d6ec584e"),
                                        data_types = "bugs",
                                        load = TRUE)
sample_experiments
```

## Combining Samples

Once the files have been loaded as SummarizedExperiment objects, matching assays
from multiple samples can be merged together with `mergeExperiments`.

```{r}
merged_experiment <- mergeExperiments(sample_experiments)
merged_experiment
```
