---
title: "Piecewise Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Piecewise Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r chunk_set, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r general_setup, include=FALSE}
library(parkinsonsMetagenomicData)
library(dplyr)
library(DBI)
```

# Piecewise Workflow

Use this vignette when you need finer control than the `First 15 Minutes`
or `Full Workflow` paths provide. If your goal is a quick first retrieval,
start with `First 15 Minutes`. If you want a complete guided wrapper-based
workflow, use `Full Workflow`.

In the piecewise workflow, we use the functions `accessParquetData` and
`loadParquetData` separately to have a little more control over our data. First,
`accessParquetData` sets up a DuckDB connection to the parquet files that are
either stored locally or hosted in a remote repo (in this case, Hugging Face).
`loadParquetData` then takes an argument for data type as well as any filters
and loads the requested data into R as a Tree Summarized Experiment. We can even
use `dplyr` functions to customize the SQL query.

## File Selection

We can use the same resources of `get_repo_info()` and `get_hf_parquet_urls()`
to determine our files of interest and `data_type` value. As before, we can
download any files and alternatively supply them as local files.

We then run `accessParquetData`, which creates the DuckDB connection and sets
up DuckDB "VIEW" objects for each available file, whether remotely hosted or
locally stored. It returns a DuckDB connection object. If you are using a remote
repo, the `data_types` argument can be left blank to access all available files,
or a smaller number can be provided. DBI functions can then be used to see which
views are now available.

```{r access_wrapper}
con <- accessParquetData(dbdir = ":memory:",
                        repo = "waldronlab/metagenomics_mac",
                        local_files = NULL,
                        data_types = "relative_abundance")
DBI::dbListTables(con)
```

Supplying local files would look much as expected, and the "data_types" argument
need not be supplied:

```{r local_access, eval=interactive()}
local_files <- c("/path/to/relative_abundance_clade_name_species.parquet",
                "/path/to/relative_abundance_uuid.parquet")

con <- accessParquetData(dbdir = ":memory:",
                        repo = NULL,
                        local_files = local_files,
                        data_types = NULL)
```

## Selecting Samples

The sample selection process is the same as in the standard workflow, with the
exception that we are only passing the UUID column to the loading function. Here
we will create the same table as earlier, then pull just the UUIDS. This is
again an optional step, as you may want to get data across all samples for
meta-analyses.

```{r re_pick_metadata}
# Load sampleMetadata
if (!exists("sampleMetadata", envir = environment())) {
    data("sampleMetadata", package = "parkinsonsMetagenomicData", envir = environment())
}

sample_table <- sampleMetadata |>
    filter(study_name == "ZhangM_2023") |>
    select(where(~ !any(is.na(.x))))
```
```{r re_show_metadata, echo=FALSE}
sample_table
```

```{r pull_uuids}
selected_uuids <- sample_table$uuid
selected_uuids
```

## Selecting Features

Our feature table is treated similarly, in that we are no longer providing the
whole table. Instead, we construct our filtering arguments as a named list. The
element name is the column to filter by, and the element values will be exact
matches. If this is our previous table:

```{r re_re_filter_feature_table}
clade_name_ref <- load_ref("clade_name_ref")
feature_table <- clade_name_ref %>%
    filter(grepl("Faecalibacterium", clade_name_genus)) %>%
    filter(!is.na(clade_name_species)) %>%
    filter(!is.na(clade_name_terminal))
```
```{r show_new_new_feature_table, echo=FALSE}
feature_table
```

We would set up the filters like so. These two columns are the minimum required
to produce the same result as our prior example.

```{r filter_vals}
filter_values <- list(clade_name_species = unique(feature_table$clade_name_species),
                    clade_name_terminal = unique(feature_table$clade_name_terminal))
filter_values
```

Here is also where we supply our UUIDs.

```{r filter_vals_with_uuid}
filter_values <- c(list(uuid = selected_uuids), filter_values)
filter_values
```

## Loading into R

We then provide our list of filtering arguments to `loadParquetData` along with
the database connection object and the data type we are accessing and receive a
Tree Summarized Experiment object. Loading the data may take some time depending
on the file type and queries.

```{r load_wrapper}
basic_experiment <- loadParquetData(con = con,
                                    data_type = "relative_abundance",
                                    filter_values = filter_values,
                                    custom_view = NULL,
                                    include_empty_samples = TRUE,
                                    dry_run = FALSE)
basic_experiment
```

Performing a dry run in order to examine the SQL query would look much as
before:

```{r query_load_wrapper}
query_only <- loadParquetData(con = con,
                                data_type = "relative_abundance",
                                filter_values = filter_values,
                                custom_view = NULL,
                                include_empty_samples = TRUE,
                                dry_run = TRUE)
dplyr::show_query(query_only)
```

## Optional View Customization

Additionally, it is possible to use `dplyr` to preview the data view of interest
and perform more advanced functions on the data prior to using `loadParquetData`
or the `collect()` function to load it into R. Here we use the `tbl()` function
to access the data view of choice, and pipe it into a `filter()` call to select
only rows that have a value in the "additional_species" column. We can then save
these calls to a variable prior to loading them into R, and supply this to
`loadParquetData` either instead of or alongside our other standard filtering
arguments.

```{r custom_view}
custom_filter <- tbl(con, "relative_abundance_uuid") %>%
    filter(!is.na(additional_species))
```

We can see a preview of the data by calling the saved view:

```{r view_custom_view}
custom_filter
```

Or show the query prior to running:

```{r query_custom_view}
dplyr::show_query(custom_filter)
```

We then simply provide that custom view to `loadParquetData()`.

```{r load_custom}
custom_experiment <- loadParquetData(con = con,
                                    data_type = "relative_abundance",
                                    filter_values = filter_values,
                                    custom_view = custom_filter,
                                    include_empty_samples = TRUE,
                                    dry_run = FALSE)
custom_experiment
```

```{r session_info}
sessionInfo()
```
