---
title: "Piecewise Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Piecewise Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r chunk_set, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r general_setup, include=FALSE}
library(parkinsonsMetagenomicData)
library(dplyr)
library(DBI)
library(DT)
```

# Piecewise Workflow

## When to Use This Approach

The "piecewise" name refers to splitting data retrieval into separate steps
(accessParquetData → custom dplyr operations → loadParquetData) rather than
using the single-step returnSamples() wrapper.

Use a piecewise workflow when you need:

1. **Direct SQL control**: Write custom dplyr queries on the DuckDB connection before loading data
2. **Multiple queries from one connection**: Query different data types efficiently and without reconnecting, such as to determine association between a species and a pathway
3. **Complex filtering**: Apply advanced dplyr operations beyond simple column matching
4. **Memory-efficient workflows**: Preview data size with `dplyr::show_query()` before loading
5. **Debugging queries**: Inspect the SQL generated at each step

If you just want to retrieve data quickly, use `First 15 Minutes` or `Full Workflow` instead. Those vignettes use the wrapper function `returnSamples()` which handles the connection and loading steps for you.

## How It Works

The piecewise workflow splits data retrieval into two steps:

1. **`accessParquetData()`** - Creates a DuckDB connection and sets up VIEW objects for parquet files (remote or local)
2. **`loadParquetData()`** - Queries those views with your filters and loads data as a TreeSummarizedExperiment

Between these steps, you can use `dplyr` to inspect, modify, and optimize queries before executing them.

## Discovering Available Data

Before selecting files, explore what data types and reference tables are available:

```{r discover_data_types}
available_data <- get_hf_parquet_urls(repo_name = "waldronlab/metagenomics_mac")
```

```{r show_data_types, echo=FALSE}
available_data |>
    select(data_type, tool, description) |>
    distinct() |>
    datatable(
        options = list(pageLength = 10),
        filter = 'top',
        caption = "Available data types",
        extensions = "Responsive"
    )
```

```{r discover_ref_tables}
ref_info <- get_ref_info()
```

```{r show_ref_tables, echo=FALSE}
ref_info |>
    datatable(
        options = list(pageLength = 10),
        filter = 'top',
        caption = "Available reference tables",
        extensions = "Responsive"
    )
```

## File Selection

Now that we know what's available, we can select our files of interest. The `data_type` value corresponds to the parquet files we want to access.

We then run `accessParquetData`, which creates the DuckDB connection and sets
up DuckDB "VIEW" objects for each available file, whether remotely hosted or
locally stored. It returns a DuckDB connection object. If you are using a remote
repo, the `data_types` argument can be left blank to access all available files,
or a smaller number can be provided. DBI functions can then be used to see which
views are now available.

```{r access_wrapper}
con <- accessParquetData(dbdir = ":memory:",
                        repo = "waldronlab/metagenomics_mac",
                        local_files = NULL,
                        data_types = "relative_abundance")
DBI::dbListTables(con)
```

Supplying local files would look much as expected, and the "data_types" argument
need not be supplied:

```{r local_access, eval=interactive()}
local_files <- c("/path/to/relative_abundance_clade_name_species.parquet",
                "/path/to/relative_abundance_uuid.parquet")

con <- accessParquetData(dbdir = ":memory:",
                        repo = NULL,
                        local_files = local_files,
                        data_types = NULL)
```

## Selecting Samples

The sample selection process is the same as in the standard workflow, with the
exception that we are only passing the UUID column to the loading function. Here
we will create the same table as earlier, then pull just the UUIDS. This is
again an optional step, as you may want to get data across all samples for
meta-analyses.

```{r re_pick_metadata}
data("sampleMetadata", package = "parkinsonsMetagenomicData")
sample_table <- sampleMetadata |>
    filter(study_name == "ZhangM_2023") |>
    select(where(~ !any(is.na(.x))))
```
```{r re_show_metadata, echo=FALSE}
sample_table |>
    datatable(extensions = "Responsive")
```

```{r pull_uuids}
selected_uuids <- sample_table$uuid
selected_uuids
```

## Selecting Features

Our feature table is treated similarly, in that we are no longer providing the
whole table. Instead, we construct our filtering arguments as a named list. The
element name is the column to filter by, and the element values will be exact
matches. If this is our previous table:

```{r re_re_filter_feature_table}
clade_name_ref <- load_ref("clade_name_ref")
feature_table <- clade_name_ref %>%
    filter(grepl("Faecalibacterium", clade_name_genus)) %>%
    filter(!is.na(clade_name_species)) %>%
    filter(!is.na(clade_name_terminal))
```
```{r show_new_new_feature_table, echo=FALSE}
feature_table |>
    datatable(extensions = "Responsive")
```

We would set up the filters like so. These two columns are the minimum required
to produce the same result as our prior example.

```{r filter_vals}
filter_values <- list(clade_name_species = unique(feature_table$clade_name_species),
                    clade_name_terminal = unique(feature_table$clade_name_terminal))
```

Here is also where we supply our UUIDs to specify samples.

```{r filter_vals_with_uuid}
filter_values[["uuid"]] <- selected_uuids
filter_values
```

## Loading into R

We then provide our list of filtering arguments to `loadParquetData` along with
the database connection object and the data type we are accessing and receive a
Tree Summarized Experiment object. Loading the data may take some time depending
on the file type and queries.

```{r load_wrapper}
basic_experiment <- loadParquetData(con = con,
                                    data_type = "relative_abundance",
                                    filter_values = filter_values,
                                    custom_view = NULL,
                                    include_empty_samples = TRUE,
                                    dry_run = FALSE)
basic_experiment
```

Performing a dry run in order to examine the SQL query would look much as
before:

```{r query_load_wrapper}
query_only <- loadParquetData(con = con,
                                data_type = "relative_abundance",
                                filter_values = filter_values,
                                custom_view = NULL,
                                include_empty_samples = TRUE,
                                dry_run = TRUE)
dplyr::show_query(query_only)
```

## Example 2: Pathway Abundance

The piecewise workflow works identically for functional data types. Here's an example with pathway abundance:

```{r pathway_example}
# Connect to pathway data
con_pathways <- accessParquetData(dbdir = ":memory:",
                                   repo = "waldronlab/metagenomics_mac",
                                   data_types = "pathabundance_unstratified")

DBI::dbListTables(con_pathways)

# Load pathway reference
pathway_ref <- load_ref("pathway_ref")

# Filter for butyrate biosynthesis pathways
butyrate_pathways <- pathway_ref |>
    filter(grepl("butanoate|butyrat", pathway, ignore.case = TRUE)) |>
    filter(!grepl("\\|", pathway)) |>  # Exclude stratified versions
    select(pathway)

# Set up filter values (same samples as before)
pathway_filter_values <- list(
    uuid = selected_uuids,
    pathway = butyrate_pathways$pathway
)

# Load pathway data
pathway_experiment <- loadParquetData(
    con = con_pathways,
    data_type = "pathabundance_unstratified",
    filter_values = pathway_filter_values,
    include_empty_samples = TRUE,
    dry_run = FALSE
)

pathway_experiment
```

This demonstrates the flexibility of the piecewise approach: you can create multiple connections to different data types and query them with the same workflow.

## Custom View: Advanced dplyr Operations

The real power of the piecewise workflow is using `dplyr` to manipulate DuckDB views before loading data. You can access any view with `tbl()` and apply dplyr operations.

### Example 1: Filter on Additional Columns

```{r custom_view}
# Access the view and add custom filtering
custom_filter <- tbl(con, "relative_abundance_uuid") %>%
    filter(!is.na(additional_species))

# Preview the data (lazy query, not executed yet)
custom_filter
```

View the generated SQL:

```{r query_custom_view}
dplyr::show_query(custom_filter)
```

Load the data with the custom view:

```{r load_custom}
custom_experiment <- loadParquetData(con = con,
                                    data_type = "relative_abundance",
                                    filter_values = filter_values,
                                    custom_view = custom_filter,
                                    include_empty_samples = TRUE,
                                    dry_run = FALSE)
custom_experiment
```

### Example 2: Aggregate Before Loading

You can perform aggregations in DuckDB before bringing data into R. This is memory-efficient for large datasets:

```{r custom_aggregation, eval=interactive()}
# Calculate summary statistics for Akkermansia muciniphila across all samples
# First check what Akkermansia species are available
akk_species <- load_ref("clade_name_ref") %>%
    filter(grepl("Akkermansia", clade_name_genus)) %>%
    pull(clade_name_species) %>%
    unique()

akk_species  # Should show "Akkermansia_muciniphila" if available

# Aggregate abundance data in DuckDB before loading into R
aggregated_view <- tbl(con, "relative_abundance_uuid") %>%
    filter(clade_name_species == "Akkermansia_muciniphila") %>%
    group_by(clade_name_species) %>%
    summarize(
        mean_abundance = mean(relative_abundance, na.rm = TRUE),
        median_abundance = median(relative_abundance, na.rm = TRUE),
        max_abundance = max(relative_abundance, na.rm = TRUE),
        n_samples_detected = sum(relative_abundance > 0, na.rm = TRUE),
        .groups = "drop"
    )

# Show the SQL that will be executed
show_query(aggregated_view)

# Execute and collect the aggregated result (small, just summary stats)
aggregated_data <- collect(aggregated_view)
aggregated_data
```

This aggregation happens in DuckDB before loading into R, making it much more memory-efficient than loading all individual sample abundances first and then aggregating in R.

### Example 3: Combine Data from Multiple Connections

You can query different data types separately and combine them in R. This example pairs taxonomic and pathway data:

```{r custom_join}
# Get species present in our samples (aggregate to species level)
# Aggregating here avoids duplicate rows from subspecies/strains
taxa_data <- tbl(con, "relative_abundance_uuid") %>%
    filter(uuid %in% selected_uuids) %>%
    filter(clade_name_species %in% !!unique(feature_table$clade_name_species)) %>%
    group_by(uuid, clade_name_species) %>%
    summarize(relative_abundance = sum(relative_abundance, na.rm = TRUE), .groups = "drop") %>%
    collect()

# Get butyrate pathway data for same samples
pathway_data <- tbl(con_pathways, "pathabundance_unstratified_uuid") %>%
    filter(uuid %in% selected_uuids) %>%
    filter(grepl("butanoate|butyrat", pathway, ignore.case = TRUE)) %>%
    filter(!grepl("\\|", pathway)) %>%  # Exclude stratified
    group_by(uuid) %>%
    summarize(total_butyrate_pathways = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
    collect()

# Join in R on uuid (sample ID)
joined_data <- taxa_data %>%
    inner_join(pathway_data, by = "uuid") %>%
    arrange(desc(relative_abundance))

```

```{r showjoined_data, echo=FALSE}
joined_data |>
    datatable(extensions = "Responsive")
```

This demonstrates querying different data types separately (aggregating in DuckDB before loading) then combining in R. The joined data pairs each sample's Faecalibacterium abundance with its total butyrate pathway abundance, enabling correlation analysis to test hypotheses like: "Do samples with higher Faecalibacterium also have higher butyrate biosynthesis capacity?"

```{r session_info}
sessionInfo()
```
