---
title: "Working with Large Parquet Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Large Parquet Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r chunk_set, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r general_setup, include=FALSE}
library(parkinsonsMetagenomicData)
library(dplyr)
library(DBI)
library(DT)
```

# Working with Large Parquet Files

This vignette focuses on strategies for working with the largest data types in 
the package, particularly `genefamilies_stratified` and similar large files. 
While these files work the same way as smaller data types, they require careful 
query strategies to avoid rate limiting and performance issues.

## Understanding Parquet File Sorting

Each data type has parquet files sorted by specific columns. This sorting is 
crucial for query performance when accessing remote files.

For example, `genefamilies_stratified` has two parquet files:
- `genefamilies_stratified_uuid.parquet` - sorted by `uuid` column
- `genefamilies_stratified_gene_family_uniref.parquet` - sorted by 
  `gene_family_uniref` column

**Key principle**: When querying remote parquet files, always filter on sorted 
columns first to minimize data transfer and avoid rate limiting.

## The Sorted Column Strategy

### Why This Matters

When you filter on a **sorted column**, DuckDB can efficiently skip large 
portions of the file without reading them. When you filter on a 
**non-sorted column**, DuckDB must scan the entire file remotely, which:

1. Takes much longer
2. Transfers more data
3. Can trigger Hugging Face rate limits (HTTP 429 errors)
4. May cause the query to hang or fail

### Two-Stage Filtering Approach

For large files, use this two-stage strategy:

**Stage 1: Remote filtering on sorted columns**
- Filter by sorted columns (e.g., `uuid`, `gene_family_uniref`) when querying 
  the remote parquet files
- This creates a smaller TreeSummarizedExperiment with only relevant data

**Stage 2: Local filtering on non-sorted columns**
- Once data is in R, apply additional filters on non-sorted columns 
  (e.g., `gene_family_species`)
- Use standard R/dplyr operations on the TreeSummarizedExperiment object

### Alternative: Download Files Locally

For repeated queries or very large extractions, download the parquet files 
locally. This completely eliminates rate limits and allows filtering on any 
column. See the `Full Workflow` vignette for a working example of local file 
usage with `system.file()` and the inst/extdata example files.

## Practical Example: Querying genefamilies_stratified

The examples below demonstrate the sorted column strategy with 
`genefamilies_stratified`, one of the largest data types. These queries use 
`accessParquetData()` and `loadParquetData()` for fine-grained control.

### Resource Considerations

Large file operations may require:
- Adequate disk space if using a persistent DuckDB database file (specify with 
  `dbdir` parameter)
- Sufficient RAM for in-memory operations (default `:memory:`)
- For very large extractions, consider running in a high-memory environment or 
  cloud instance

### Example Scenario: Mouse Microbiome Gene Families

This example retrieves specific gene families from a mouse study, demonstrating 
the sorted column strategy:

```{r demo_large_queries}
# Establish connection to genefamilies_stratified remote files
con_gs <- accessParquetData(data_types = "genefamilies_stratified")

# Target: Gene families specific to different mouse groups
# - UniRef90_T4BVE4: present only in SPF mice
# - UniRef90_A0A1B1SA57: present only in WildR mice

# Load study metadata
data("sampleMetadata", package = "parkinsonsMetagenomicData")
selected_samples <- sampleMetadata |>
    filter(study_name == "MazmanianS_DumitrescuDG") |>
    select(where(~ !any(is.na(.x))))

# STRATEGY: Pre-filter sample IDs using metadata before querying parquet files
# This minimizes the data transfer from remote files

wildr_ids <- selected_samples |>
    filter(MazmanianS_DumitrescuDG_uncurated_donor_microbiome_type == "WildR") |>
    pull(uuid)

spf_ids <- selected_samples |>
    filter(MazmanianS_DumitrescuDG_uncurated_donor_microbiome_type == "SPF") |>
    pull(uuid)

# Query 1: SPF-specific gene family with subset of samples
# Filters on SORTED columns: gene_family_uniref and uuid
spf_ex <- loadParquetData(con_gs, "genefamilies_stratified",
        filter_values = list(gene_family_uniref = "UniRef90_T4BVE4",
                            uuid = spf_ids))
spf_ex

# Query 2: WildR-specific gene family with subset of samples
wildr_ex <- loadParquetData(con_gs, "genefamilies_stratified",
    filter_values = list(gene_family_uniref = "UniRef90_A0A1B1SA57",
                        uuid = wildr_ids))
wildr_ex

# Query 3: Multiple gene families across all samples in study
# This works but transfers more data (all samples vs. filtered subset)
all_ex <- loadParquetData(con_gs, "genefamilies_stratified",
                filter_values = list(gene_family_uniref = c("UniRef90_T4BVE4",
                                                        "UniRef90_A0A1B1SA57"),
                                    uuid = selected_samples$uuid))
all_ex
```

### Key Takeaways

1. **Always filter on sorted columns** when querying remote large files
2. **Pre-filter sample IDs** using metadata before querying parquet files
3. **Balance specificity vs. breadth**: More specific queries 
  (fewer samples/features) are faster
4. For filtering on non-sorted columns (e.g., `gene_family_species`), retrieve 
  data first, then filter the TreeSummarizedExperiment in R

## When to Download Locally

Consider downloading parquet files if:
- You need to query the same large files repeatedly
- You need to filter on non-sorted columns extensively
- You're working in an environment with fast local disk but slow/unreliable internet
- You're hitting rate limits despite using sorted column strategies

See the `Full Workflow` vignette for local file examples.

```{r session_info}
sessionInfo()
```
